# Exercises for 'Mapping over functions' chapter 

```{r}
#| include: false
# if it is available, run the setup script that tells quarto to round all df/tibble outputs to three decimal places
if(file.exists("../_setup.R")){source("../_setup.R")}
```

These exercises accompany the [mapping chapter](../chapters/08_mapping.qmd).

You can complete these exercises in your local version of the .qmd file. Either download a copy of the whole book from github (see introduction), or download this .qmd using the download button on the top right of the page.

Write code for each of the following. Remember that you can write pseudocode first if it helps.

```{r}
#| include: false
# dependencies
library(tibble)
library(dplyr)
library(tidyr)
library(purrr)
library(furrr)
library(ggplot2)

set.seed(42)

generate_data <- function(n_per_condition,
                          mean_control,
                          mean_intervention,
                          sd) {

  data_control <-
    tibble(condition = "control",
           score = rnorm(n = n_per_condition, mean = mean_control, sd = sd))

  data_intervention <-
    tibble(condition = "intervention",
           score = rnorm(n = n_per_condition, mean = mean_intervention, sd = sd))

  bind_rows(data_control, data_intervention)
}

analyse_data <- function(data) {
  res_t_test <- t.test(formula = score ~ condition,
                       data = data,
                       var.equal = TRUE,
                       alternative = "two.sided")

  tibble(p = res_t_test$p.value)
}
```

## Warm-up (brief)

### Repeat a simulation with `map()`

Use the `generate_data()` and `analyse_data()` helper functions defined above.

Create `iterations <- 1:20`, then use `map()` to run the same simulation 20 times with:

-   `n_per_condition = 50`
-   `mean_control = 0`
-   `mean_intervention = 0.5`
-   `sd = 1`

Inspect the output. What object type did `map()` return?

```{r}



```

### Return a numeric vector with `map_dbl()`

Repeat the previous exercise using `map_dbl()` to extract only the `p` values into a numeric vector. Then calculate the proportion of p-values less than 0.05.

```{r}



```

## Parameter grids + `pmap()` (our main focus)

### Create a parameter grid with `expand_grid()`

Use `expand_grid()` to create a parameter grid where:

-   `iteration = 1:200`
-   `n_per_condition = c(20, 60)`
-   `mean_intervention = c(0, 0.4, 0.8)`
-   `mean_control = 0`
-   `sd = c(0.5, 1)`

After creating the grid:

1.  Check the total number of rows.
2.  Use `distinct()` to verify the unique combinations of design parameters (excluding `iteration`).

```{r}



```

### Generate data row-wise with `pmap()`

Using your parameter grid, create a new object where each row has a list-column called `generated_data`, produced with `pmap()` and `generate_data()`.

```{r}



```

### Analyse each generated dataset and unnest

Starting from your object with the `generated_data` list-column:

1.  Use `map()` to apply `analyse_data()` to each generated dataset.
2.  Store the output in a `results` list-column.
3.  Use `unnest()` to create one tidy simulation-results tibble.

```{r}



```

## Write reusable wrappers

### Write a simulation wrapper using `expand_grid()` + `pmap()`

Write a function called `run_simulation_pmap()` with arguments:

-   `n_iterations`
-   `n_per_condition`
-   `mean_intervention`
-   `mean_control = 0`
-   `sd = 1`
-   `keep_data = FALSE`

Function requirements:

1.  Build a parameter grid with `expand_grid()`.
2.  Generate data using `pmap()`.
3.  Analyse data using `map()` + `analyse_data()`.
4.  Return an unnested tibble.
5.  If `keep_data = FALSE`, drop the `generated_data` list-column before returning.

```{r}



```

## A whole new simulation, from scratch 

### Build a new simulation with binary outcomes using `expand_grid()` + `pmap()`

In this final exercise, write a full simulation workflow for a **different** data-generating process (binary outcomes instead of continuous scores), covering steps 1 through 4 of a simulation:

1.  Design the experiment.
2.  Write a `generate_data()` function.
3.  Write an `analyse_data()` function.
4.  Run the workflow many times using mapping.

Use the following specification:

1.  Build a parameter grid called `experiment_parameters_binary` using `expand_grid()` with `iteration = 1:300`, `n_per_condition = c(40, 120)`, `prob_control = c(0.20, 0.35)`, and `risk_difference = c(0, 0.10)`.

2.  Add `prob_intervention` with `mutate(prob_intervention = prob_control + risk_difference)`.

3.  Write `generate_data_binary(n_per_condition, prob_control, prob_intervention)`. The function should return a tibble with columns `condition` (`"control"` / `"intervention"`) and `outcome` (0/1, generated with `rbinom()`).

4.  Write `analyse_data_binary(data)` that returns a one-row tibble with columns `p` (from `prop.test()` comparing intervention vs control) and `risk_difference_observed` (observed mean outcome in intervention minus control).

5.  Run the simulation by using `pmap()` to generate a data for each row of `experiment_parameters_binary` (making sure the generated data are stored in the output), then `map()` to apply `analyse_data_binary()`, then `unnest()` into one tibble.

6.  Do some basic checks: confirm the number of output rows equals the number of rows in the parameter grid, and verify all `p` values are between 0 and 1.

```{r}



```


## Optional extension: parallel mapping with `{furrr}`

### Use `future_pmap()` and `future_map()`

Rewrite your pipeline using `future_pmap()` and `future_map()`. Use `plan(multisession, workers = parallelly::availableCores())` and `furrr_options(seed = TRUE)` to keep results reproducible. Also add `slice_sample(prop = 1)` to the experiment grid before mapping over it, so that the computationally expensive iterations are load balanced evenly across the workers.

```{r}



```
