# Exercises for 'p-hacking' chapter

```{r}
#| include: false
# if it is available, run the setup script that tells quarto to round all df/tibble outputs to three decimal places
if(file.exists("../_setup.R")){source("../_setup.R")}
```

```{r}
# dependencies
library(tidyr)
library(dplyr)
library(readr)
library(forcats)
library(stringr)
library(ggplot2)
library(scales)
library(patchwork)
library(tibble)
library(purrr)
library(furrr)
library(simhelpers)
library(faux)
library(janitor)

# set up parallelization
# use availableCores() to define the number of cores, in case the simulation is being
# run on a HPC that might have more physical cores than are available to the session
plan(multisession, workers = parallelly::availableCores())
```

## Basis simulation

Same as in the chapter: selective reporting across two studies.

```{r}

generate_data <- function(n_per_condition,
                          mean_control,
                          mean_intervention,
                          sd) {

  data_control <-
    tibble(condition = "control",
           score = rnorm(n = n_per_condition, mean = mean_control, sd = sd))

  data_intervention <-
    tibble(condition = "intervention",
           score = rnorm(n = n_per_condition, mean = mean_intervention, sd = sd))

  data_combined <- bind_rows(data_control,
                             data_intervention)

  return(data_combined)
}


# define data analysis function ----
analyze <- function(data) {

  students_ttest <- t.test(formula = score ~ condition,
                           data = data,
                           var.equal = TRUE,
                           alternative = "two.sided")

  res <- tibble(p = students_ttest$p.value)

  return(res)
}


# define experiment parameters ----
experiment_parameters <- expand_grid(
  n_per_condition = 100,
  mean_control = 0,
  mean_intervention = 0,
  sd = 1,
  iteration = 1:1000
)


# run simulation ----
# results are cached to disk: the simulation is only run the first time, and read back
# in on subsequent renders. delete the .rds file if you change the simulation code.
dir.create("results", showWarnings = FALSE)

if(file.exists("results/simulation_p_hacking_exercises_two_studies.rds")){

  simulation <- read_rds("results/simulation_p_hacking_exercises_two_studies.rds")

} else {

  set.seed(42)

  simulation <-
    # using the experiment parameters
    experiment_parameters |>

    # shuffle the rows of the grid to load balance the computationally expensive
    # iterations evenly across the workers
    slice_sample(prop = 1) |>

    # generate data using the data generating function and the parameters relevant to data generation
    mutate(generated_data_study1 = future_pmap(.l = list(n_per_condition,
                                                         mean_control,
                                                         mean_intervention,
                                                         sd),
                                               .f = generate_data,
                                               .progress = TRUE,
                                               .options = furrr_options(seed = TRUE))) |>
    mutate(generated_data_study2 = future_pmap(.l = list(n_per_condition,
                                                         mean_control,
                                                         mean_intervention,
                                                         sd),
                                               .f = generate_data,
                                               .progress = TRUE,
                                               .options = furrr_options(seed = TRUE))) |>

    # apply the analysis function to the generated data using the parameters relevant to analysis
    mutate(results_study1 = future_pmap(.l = list(data = generated_data_study1),
                                        .f = analyze,
                                        .progress = TRUE,
                                        .options = furrr_options(seed = TRUE))) |>
    mutate(results_study2 = future_pmap(.l = list(data = generated_data_study2),
                                        .f = analyze,
                                        .progress = TRUE,
                                        .options = furrr_options(seed = TRUE))) |>

    # undo the shuffle so the rows are back in the order of the experiment grid
    arrange(iteration) |>

    # the generated data sets are not needed after they have been analyzed, and keeping
    # them would make the saved file enormous
    select(-generated_data_study1, -generated_data_study2)

  write_rds(x = simulation,
            file = "results/simulation_p_hacking_exercises_two_studies.rds",
            compress = "gz")

}


# summarise simulation results over the iterations ----
simulation_summary <- simulation |>
  # unnest and rename
  unnest(results_study1) |>
  rename(p_study1 = p) |>
  unnest(results_study2) |>
  rename(p_study2 = p) |>
  # simulate flexible reporting
  mutate(p_hacked = ifelse(p_study1 < .05, p_study1, p_study2)) |>
  # summarize. rather than calculating the proportion of significant results by hand,
  # use {simhelpers}'s dedicated function for the empirical rejection rate.
  # calc_rejection() also returns the number of iterations and the Monte Carlo Standard
  # Error of the rate; select() keeps and renames just the rate itself
  reframe(calc_rejection(data     = pick(everything()),
                         p_values = p_study1,
                         alpha    = 0.05) |>
            select(prop_sig_study1 = rej_rate),
          calc_rejection(data     = pick(everything()),
                         p_values = p_study2,
                         alpha    = 0.05) |>
            select(prop_sig_study2 = rej_rate),
          calc_rejection(data     = pick(everything()),
                         p_values = p_hacked,
                         alpha    = 0.05) |>
            select(prop_sig_hacked = rej_rate)) |>
  pivot_longer(cols = everything(),
               names_to = "Source",
               values_to = "proportion_significant") |>
  mutate(Source = str_remove(Source, "prop_sig_"))

```

### Results

```{r}

simulation_summary |>
  mutate_if(is.numeric, round_half_up, digits = 2)

```

The false positive rate for two studies is \~10%, because each study has a 5% false positive rate.

## Exercises

For each of the exercises below, create a copy of the simulation above. Modify it appropriately to answer the exercise.

### Extend the simulation to four studies

Extend the simulation so that four studies are run. What do you expect the false positive rate to be? What do you find?

```{r}



```

### Solution

```{r}

generate_data <- function(n_per_condition,
                          mean_control,
                          mean_intervention,
                          sd) {

  data_control <-
    tibble(condition = "control",
           score = rnorm(n = n_per_condition, mean = mean_control, sd = sd))

  data_intervention <-
    tibble(condition = "intervention",
           score = rnorm(n = n_per_condition, mean = mean_intervention, sd = sd))

  data_combined <- bind_rows(data_control,
                             data_intervention)

  return(data_combined)
}


# define data analysis function ----
analyze <- function(data) {

  students_ttest <- t.test(formula = score ~ condition,
                           data = data,
                           var.equal = TRUE,
                           alternative = "two.sided")

  res <- tibble(p = students_ttest$p.value)

  return(res)
}


# define experiment parameters ----
experiment_parameters <- expand_grid(
  n_per_condition = 100,
  mean_control = 0,
  mean_intervention = 0,
  sd = 1,
  iteration = 1:1000
)


# run simulation ----
# results are cached to disk: the simulation is only run the first time, and read back
# in on subsequent renders. delete the .rds file if you change the simulation code.
dir.create("results", showWarnings = FALSE)

if(file.exists("results/simulation_p_hacking_exercises_four_studies.rds")){

  simulation <- read_rds("results/simulation_p_hacking_exercises_four_studies.rds")

} else {

  set.seed(42)

  simulation <-
    # using the experiment parameters
    experiment_parameters |>

    # shuffle the rows of the grid to load balance the computationally expensive
    # iterations evenly across the workers
    slice_sample(prop = 1) |>

    # generate data using the data generating function and the parameters relevant to data generation
    mutate(generated_data_study1 = future_pmap(.l = list(n_per_condition = n_per_condition,
                                                         mean_control = mean_control,
                                                         mean_intervention = mean_intervention,
                                                         sd = sd),
                                               .f = generate_data,
                                               .progress = TRUE,
                                               .options = furrr_options(seed = TRUE))) |>
    mutate(generated_data_study2 = future_pmap(.l = list(n_per_condition = n_per_condition,
                                                         mean_control = mean_control,
                                                         mean_intervention = mean_intervention,
                                                         sd = sd),
                                               .f = generate_data,
                                               .progress = TRUE,
                                               .options = furrr_options(seed = TRUE))) |>
    mutate(generated_data_study3 = future_pmap(.l = list(n_per_condition = n_per_condition,
                                                         mean_control = mean_control,
                                                         mean_intervention = mean_intervention,
                                                         sd = sd),
                                               .f = generate_data,
                                               .progress = TRUE,
                                               .options = furrr_options(seed = TRUE))) |>
    mutate(generated_data_study4 = future_pmap(.l = list(n_per_condition = n_per_condition,
                                                         mean_control = mean_control,
                                                         mean_intervention = mean_intervention,
                                                         sd = sd),
                                               .f = generate_data,
                                               .progress = TRUE,
                                               .options = furrr_options(seed = TRUE))) |>

    # apply the analysis function to the generated data using the parameters relevant to analysis
    mutate(results_study1 = future_pmap(.l = list(data = generated_data_study1),
                                        analyze,
                                        .progress = TRUE,
                                        .options = furrr_options(seed = TRUE))) |>
    mutate(results_study2 = future_pmap(.l = list(data = generated_data_study2),
                                        analyze,
                                        .progress = TRUE,
                                        .options = furrr_options(seed = TRUE))) |>
    mutate(results_study3 = future_pmap(.l = list(data = generated_data_study3),
                                        analyze,
                                        .progress = TRUE,
                                        .options = furrr_options(seed = TRUE))) |>
    mutate(results_study4 = future_pmap(.l = list(data = generated_data_study4),
                                        analyze,
                                        .progress = TRUE,
                                        .options = furrr_options(seed = TRUE))) |>

    # undo the shuffle so the rows are back in the order of the experiment grid
    arrange(iteration) |>

    # the generated data sets are not needed after they have been analyzed, and keeping
    # them would make the saved file enormous
    select(-starts_with("generated_data_study"))

  write_rds(x = simulation,
            file = "results/simulation_p_hacking_exercises_four_studies.rds",
            compress = "gz")

}


# summarise simulation results over the iterations ----
simulation_summary <- simulation |>
  # unnest and rename
  unnest(results_study1) |>
  rename(p_study1 = p) |>
  unnest(results_study2) |>
  rename(p_study2 = p) |>
  unnest(results_study3) |>
  rename(p_study3 = p) |>
  unnest(results_study4) |>
  rename(p_study4 = p) |>
  # simulate flexible reporting
  mutate(p_hacked = case_when(p_study1 < .05 ~ p_study1,
                              p_study2 < .05 ~ p_study2,
                              p_study3 < .05 ~ p_study3,
                              TRUE ~ p_study4)) |>
  # summarize using {simhelpers}'s dedicated function for the empirical rejection rate
  reframe(calc_rejection(data = pick(everything()), p_values = p_study1, alpha = 0.05) |>
            select(prop_sig_study1 = rej_rate),
          calc_rejection(data = pick(everything()), p_values = p_study2, alpha = 0.05) |>
            select(prop_sig_study2 = rej_rate),
          calc_rejection(data = pick(everything()), p_values = p_study3, alpha = 0.05) |>
            select(prop_sig_study3 = rej_rate),
          calc_rejection(data = pick(everything()), p_values = p_study4, alpha = 0.05) |>
            select(prop_sig_study4 = rej_rate),
          calc_rejection(data = pick(everything()), p_values = p_hacked, alpha = 0.05) |>
            select(prop_sig_hacked = rej_rate)) |>
  pivot_longer(cols = everything(),
               names_to = "Source",
               values_to = "proportion_significant") |>
  mutate(Source = str_remove(Source, "prop_sig_"))

simulation_summary |>
  mutate_if(is.numeric, round_half_up, digits = 2)

```

## In-class exercises

Without writing any code, design and describe the necessary components of simulations that quantify the the effect of the following p-hacking strategies on the false positive rate (i.e., empirical detection rate when true difference in means is 0)

- Flexibility in whether the DV is log transformed or not.
- Flexibility in whether to exclude DV outliers or not (e.g., >±1SD from mean).
- Flexibility calculating a median split for the DV or not.
