# Exercises

Work through these in order. If you get stuck, try using pseudocode first.

## Reproducibility warm-up

Predict the output before you run it.

```{r}

# 1) What can this return?
# rbinom(n = 10, size = 1, prob = 0)

# 2) What about prob = 1?

# 3) What would you expect the value of mean(rbinom(n = 100, size = 1, prob = 0.7)) to be close to?

```

## One-group normal DGP

Write a function `generate_data_one_group()` that takes `n`, `mean`, and `sd` and returns a tibble with columns `id` and `score`.

```{r}



```

## Two-group DGP using Cohen’s d

Write `generate_data_two_group_d()` with arguments `n_per_condition`, `mean_control`, `d`, and `sd`.

Inside the function convert `d` to a mean difference:

$$\Delta = d \times sd$$

and set `mean_intervention = mean_control + Delta`.

```{r}



```

## Binary outcomes with a treatment effect

Rewrite `generate_data_two_group_binary()` so that it returns `id`, `condition` (as a factor), and `score`.

Simulate one dataset with `prob_control = 0.20` and `prob_intervention = 0.35` and verify the observed proportions with `group_by()` and `summarise()`.

```{r}



```

## Missing data

Modify `generate_data_two_group()` so that a proportion of `score` values are set to `NA` completely at random.

Add an argument `missing_rate` between 0 and 1. Test with `missing_rate = 0`, `0.10`, and `0.40`.

```{r}



```

## Three time points

Write a function that returns three time points (`pre`, `mid`, `post`) with correlations:

- cor(pre, mid) = 0.7
- cor(mid, post) = 0.7
- cor(pre, post) = 0.5

Assume the same SD at each time point.

```{r}



```

## Scale items with `rnorm_multi()`

Using `faux::rnorm_multi()`:

- generate 8 items with inter-item correlation `r=0.3`
- add an `id` column
- compute a `total` score as a row-mean or row-sum

```{r}



```


## Pseudocode first (required)

Choose one of the functions you wrote above and write pseudocode for it (inputs, steps, outputs). Only then write or revise the R function.

## Reflection

In one paragraph: what assumptions in your DGPs are unrealistic for real psychological data, and which one would you most like to relax later?

