# Exercises

These exercises accompany the [Data Generating Process functions chapter](../chapters/4_data_generating_process_functions.qmd).

You can complete these exercises in your local version of the .qmd file. Either download a copy of the whole book from github (see introduction), or download this .qmd using the download button on the top right of the page.

Write functions for each of the following. Remember that you can write pseudocode first if it helps.

## Normally distributed data

### Sample data from a normal distribution and return a tibble  

Generate data for a study where the population is normally distributed, with defined population mean and SD, and there is a single group.

Use `rnorm()`. 

```{r}





```

### Sample data from a normal distribution, round to the nearest whole number, and return a tibble  

This makes the data look a bit more like Likert data. Note that properly simulating realistic Likert data is harder than this.

```{r}





```

### Sample data from a normal distribution, calculate its average, and return a tibble 

```{r}





```

### Sample data from two normal distributions, return a tibble with two columns called intervention and control

This looks more like data from an RCT.

The arguments should be `mean_intervention`, `mean_control`, `sd`, and `n_per_condition`.

```{r}





```

### Sample data from two normal distributions, return a tibble in long format with a 'score' and 'condition' column

Use `bind_rows()`.

```{r}





```

### Cohen's d simulator

Write a function like the two-group RCT function we did before, but it should only take the arguments `cohens_d` and `n_per_group` (i.e., not means and SDs).

`rnorm()` can produce Cohen's d-like data when we fix the SDs to 1, one mean to 0, and the second mean to the desired population Cohen's d value.

```{r}



```

## Other probability density functions

### Generate a random number between 0 and 9

Use `runif()`. Look up the help menu if you need with `?runif()`.

```{r}





```

### Generate two random numbers between 0 and 9, return only the larger one

Use `runif()`. Look up the help menu if you need with `?runif()`.

There are multiple ways to do this. Don't reach immediately for help from an AI: think about the logic of how you might do it. Write out increasingly detailed pseudocode. There is more than one way of doing it.

```{r}





```

### Calculate the probabilty of n number of 'heads' coin flips in a row

Use `rbinom()`. Look up the help menu if you need with `?rbinom()`.

```{r}





```

### Binary outcomes with a treatment effect

Rewrite `generate_data_two_group_binary()` so that it returns `id`, `condition` (as a factor), and `score`. Use `rbinom()`.

Simulate one dataset with `prob_control = 0.20` and `prob_intervention = 0.35` and verify the observed proportions with `group_by()` and `summarise()`.

```{r}



```

### Missing data

Modify `generate_data_two_group()` so that a proportion of `score` values are set to `NA` completely at random.

Use [suggestion needed].

Add an argument `missing_rate` between 0 and 1. Test with `missing_rate = 0`, `0.10`, and `0.40`.

```{r}



```

## Multivariate normal distributions

I.e., correlated data.

## Three time points

Write a function that returns three time points (`pre`, `mid`, `post`) with correlations:

- cor(pre, mid)  = 0.7
- cor(mid, post) = 0.7
- cor(pre, post) = 0.5

Assume the same SD at each time point.

```{r}



```

## Scale items with `rnorm_multi()`

Using `faux::rnorm_multi()`:

- generate 8 items with inter-item correlation `r=0.3`
- add an `id` column
- compute a `total` score as a row-mean or row-sum

```{r}



```

## Reflection

- What features of the DGPs often assumed for psychological data are often unrealistic for real psychological data? These are the sorts of things we a) violate all the time, and yet b) usually don't understand the implications of violating (i.e., do we need to care). We can examine them in simulations in later chapters.


