---
# cache this chapter's results at the file level. the chunks below deliberately use
# simple, un-cached map()/pmap() calls to teach the mapping workflow, so caching is
# switched on here rather than complicating the example code. knitr stores each
# chunk's results in 08_mapping_cache/ and re-uses them until that chunk's code
# changes. note that the future_pmap() simulation near the end of the chapter also
# caches its results to results/simulation_mapping.rds; to force that one to re-run
# you need to delete both that .rds and the chunk's entry in 08_mapping_cache/.
execute:
  cache: true
---

# Mapping over functions <span class="badge badge-draft2">✎ Rough draft</span>

```{r}
#| include: false
# if it is available, run the setup script that tells quarto to round all df/tibble outputs to three decimal places
if(file.exists("../_setup.R")){source("../_setup.R")}
```

<!-- TODO: Fix the outputs so we just call head() and not the full output -->
<!-- TODO: Especially for the nested data -->
<!-- TODO: more explicit content on nested dataframes, visualisation of it, viewing individual datasets in the viewer, etc. -->
<!-- TODO: explain unnest() more. -->

## Overview of tutorial

Recall, once again, the five steps of Monte Carlo simulation:

1. Design an experiment
2. Generate data
3. Analyse data
4. Repeat this many times
5. Summarise over the results

In previous chapters, we have created parameter grid specifications (Step 1; `expand_grid()`), written data-generation functions (Step 2; `generate_data()`), and written analysis functions (Step 3; `analyse_data()`). We have also called `generate_data()` and `analyse_data()` together using the pipe, like: `generate_data() %>% analyse_data()`.

The question now is: how do we get that piped workflow to work with the parameter grid? In other words: how do we do this many times (i.e., Step 4) **without** copy-pasting code?

The answer is: by using the `map_*()` family of functions to map our parameter grid over the generate/analyse functions. In the tidyverse, this is usually done with the `{purrr}` package.

By the end of this chapter, you should be able to:

-   Use `map()` to repeat a simulation many times.
-   Use `map_dbl()` / `map_int()` / `map_chr()` to get clean vectors out.
-   Use `map_dfr()` to bind outputs into a tibble.
-   Use `map2()` and `pmap()` to run functions over many different parameters.
-   Understand what it means when mapping produces nested data frames (list-columns).
-   Use `future_map()` / `future_pmap()` (from `{furrr}`) when you want to parallelize.

## Recap: generate and analyse functions

First, let's set up an example that we have used in the previous chapeter: two conditions (control vs intervention) with normally distributed scores (means and SDs that we specify), and a test of differences in group means (i.e., a two-sample *t*-test).

```{r}
# dependencies
library(tidyr)
library(dplyr)
library(purrr)
library(ggplot2)
library(scales)

generate_data <- function(n_per_condition,
                          mean_control,
                          mean_intervention,
                          sd) {

  data_control <-
    tibble(condition = "control",
           score = rnorm(n = n_per_condition, mean = mean_control, sd = sd))

  data_intervention <-
    tibble(condition = "intervention",
           score = rnorm(n = n_per_condition, mean = mean_intervention, sd = sd))

  data_combined <- bind_rows(data_control, data_intervention)

  return(data_combined)
}

analyse_data <- function(data) {

  res_t_test <- t.test(formula = score ~ condition,
                       data = data,
                       var.equal = TRUE,
                       alternative = "two.sided")

  res <- tibble(p = res_t_test$p.value)

  return(res)
}
```

As we have done before, we can run the functions together using the pipe to get some output.

```{r}
results <-
  generate_data(n_per_condition = 50,
                mean_control = 0,
                mean_intervention = 0.5,
                sd = 1) %>%
  analyse_data()

results
```

What if we want to do this whole process many times? We could just copy-paste this 1000 times, run each block of code once, and save each run to a dataframe. But this is not the ideal approach, because:

-   It’s slow to write and easy to make mistakes.
-   It’s hard to scale from a few iterations to 10,000 iterations.
-   It’s hard to turn into a proper experiment (varying the parameters systematically).

So: we need a principled way to repeat the same computation. To do this, we can use the "`map()`" family of functions in R.

## The core idea of mapping

### map() repeats a function over a vector/list

`map()` takes:

-   a vector/list of inputs (.x)
-   a function (.f)

and returns a list (i.e., not a tibble) of outputs.

Let's look at a simple example. Imagine we have a vector of numbers (from 1 to 4) and, for each number, we want to add 10. We can use `map()` to do this:

```{r}
numbers <- c(1, 2, 3, 4)

map(.x = numbers, 
    .f = ~ .x + 10)
```

Notice the output is a list, even though each element is just a number. This is on purpose: returning a list is often seen as a "safe" default, because in real work you might return complicated objects (models, data frames, plots…) and lists handle these diverse objects most effectively, and prevent errors from arising. Of course, remember that in this course we want to return our simulation output as a `tibble`, not a list. We will get to this in a few sections. 

### The anonymous-function shorthand: `~`

When you look at the example above, you might be wondering: what is the `~` symbol, and what is it doing? This is referred to as the anonymous-function (lambda) shorthand in `{purrr}` formula syntax. This lets you define a short function within the `map()` family of functions. In the case above, our function was "add 10 to the input", where .x stands for the current input.

More concretely, the two lines of code below are equivalent:

Line 1:

```{r}
map(numbers, function(x) x + 10)
```

Line 2:

```{r}
map(numbers, ~ .x + 10)
```

You’ll see the \~ .x style constantly in tidyverse simulation code.

## A first simulation with map()

### Map over iterations

Ideally, simulations need an iteration index, which tells us what "round" of the simulation a given set of outputs are part of. We can often define these iterations simply, like:

```{r}
iterations <- 1:10
iterations
```

Now we map over iterations. For now, the iteration number doesn’t do anything - we just need to supply `map()` with a set of values to run across.

```{r}
simulation_results_list <-
  map(.x = iterations, 
      .f = ~ {
    generate_data(n_per_condition = 50,
                  mean_control = 0,
                  mean_intervention = 0.5,
                  sd = 1) %>%
      analyse_data()
  })

simulation_results_list
```

You should see a list of tibbles, each tibble containing one *p*-value.

If we change the number of iterations, then the number of times `map()` is called will change:

```{r}
n_iterations <- 20
iterations <- 1:n_iterations
iterations
```

```{r}
simulation_results_list <-
  map(.x = iterations, 
      .f = ~ {
    generate_data(n_per_condition = 50,
                  mean_control = 0,
                  mean_intervention = 0.5,
                  sd = 1) %>%
      analyse_data()
  })

simulation_results_list
```

### Extract one element

Elements from the list output of `map()` are accessed with \[\[ \]\]. For example, if we want the output of the first iteration, we can do this:

```{r}
simulation_results_list[[1]]
```

## Turning a list into a tibble

Even though lists are helpful for preventing errors, we ideally would prefer the output to be in a tibble format like we have previously discussed. To get a tibble as output, we can use the `map_dfr()` function, which returns dataframes as output instead of lists. If each iteration returns a tibble with the same columns (in our case, a single column called p), map_dfr() will map and then bind the outputs together into a single dataframe. We can then simply convert this dataframe back to a tibble.

```{r}

set.seed(42)

simulation_results <-
  map_dfr(.x = iterations, 
          .f = ~ {
    generate_data(n_per_condition = 50,
                  mean_control = 0,
                  mean_intervention = 0.5,
                  sd = 1) %>%
      analyse_data()
  }) %>%
  # convert back to tibble
  as_tibble()


simulation_results 
```

As you can see, now simulation_results is a single tibble, not a list.

### Add the iteration number

When we look at the nice output above from `map_dfr()`, we're missing one important piece of information: the iteration number! As we said above, we ideally want to keep this number so we know what result belongs to what step of the simulation.

We can do this by going back to the original `map()` function with some small additions. Specifically:

1.  We create a tibble with a single column, called iteration (or whatever we want), and assign this a vector of values from 1 to the number of iterations we want. For example, if we want 20 iterations:

```{r}
tibble(iteration = 1:20)
```

2.  We mutate a new column into this new tibble, called results (or whatever we want to call it). Then we run `map()` within the `mutate()` call using the iteration column and our `generate_data()` and `analyse_data()` functions, like this:

```{r}

iterations <- 1:20

simulation_results <-
  tibble(iteration = iterations) %>%
  mutate(results = map(
    .x = iteration, 
    .f = ~ {
    generate_data(n_per_condition = 50,
                  mean_control = 0,
                  mean_intervention = 0.5,
                  sd = 1) %>%
      analyse_data()
  })
  )

simulation_results
```

3.  Lastly, we can see that the results column contains a tibble per row. We simply need to extract the values from this tibble, which we can do using the `unnest()` function. The whole workflow together, then, looks like:

```{r}
simulation_results <-
  tibble(iteration = iterations) %>%
  mutate(results = map(
    .x = iteration,
    .f = ~ {
    generate_data(n_per_condition = 50,
                  mean_control = 0,
                  mean_intervention = 0.5,
                  sd = 1) %>%
      analyse_data()
  })) %>%
  unnest(results)

simulation_results
```

This "make a tibble, map into a list-column, then unnest()" pattern is one of the most fundamental and important ways to do tidy simulation.

## Other map\_\*() variants

There are several different variants of the `map()` function. You already saw one above: the `map_dfr()` function, which gave a dataframe as output instead of the list output that `map()` gives. The other `map_*()` variants follow a similar pattern - they change the type of output that comes from the `map_*()` function.

For example, imagine you want one number per iteration, rather than a dataframe or list. You might want a numeric vector of *p*-values. In this case, we could use `map_dbl()`.

### `map_dbl()` returns a numeric vector

To use map_dbl(), your function must return one numeric value.

```{r}
iterations <- 1:1000

p_values <-
  map_dbl(
    .x = iterations, 
    .f = ~ {
    generate_data(n_per_condition = 50,
                  mean_control = 0,
                  mean_intervention = 0.5,
                  sd = 1) %>%
      analyse_data() %>%
      pull(p)
  })

p_values

mean(p_values < .05)

```

Now p_values is a plain numeric vector. With this vector, we can do a quick summary, like look at the proportion of p-values less than 0.05:

```{r}
mean(p_values < .05)
```

That number is an estimate of our statistical power under these parameter settings (with potentially lots of error because we only used 10 iterations).

## Null vs alternative using mapping

Let’s examine the distribution of *p*-values under null vs alternative hypotheses using what we've learned from the `map()` functions.

Let's say we want to simulate two conditions:

-   mean_intervention = 0 (null - there is no true effect)
-   mean_intervention = 0.5 (alternative - there is a true effect)

with one of two sample sizes:

-   n_per_condition = 20
-   n_per_condition = 50

and do 2000 iterations for each combination.

This means we have four combinations in total. We could handle this using `map()`, defining each of those conditions separately, for example like:

```{r}
iterations <- 1:100 # lowered for testing; 2000 would be more appropriate 

# null results, twenty per group
simulation_results_null_twenty_per_group <-
  tibble(iteration = iterations) %>%
  mutate(results = map(
    .x = iteration, 
    .f = ~ {
    generate_data(n_per_condition = 20,
                  mean_control = 0,
                  mean_intervention = 0,
                  sd = 1) %>%
      analyse_data()
  })) %>%
  unnest(results)

# null results, fifty per group
simulation_results_null_fifty_per_group <-
  tibble(iteration = iterations) %>%
  mutate(results = map(
    .x = iteration,
    .f = ~ {
    generate_data(n_per_condition = 50,
                  mean_control = 0,
                  mean_intervention = 0,
                  sd = 1) %>%
      analyse_data()
  })) %>%
  unnest(results)

# alternative results, twenty per group
simulation_results_alternative_twenty_per_group <-
  tibble(iteration = iterations) %>%
  mutate(results = map(
    .x = iteration, 
    .f = ~ {
    generate_data(n_per_condition = 20,
                  mean_control = 0,
                  mean_intervention = 0.5,
                  sd = 1) %>%
      analyse_data()
  })) %>%
  unnest(results)

# alternative results, fifty per group
simulation_results_alternative_fifty_per_group <-
  tibble(iteration = iterations) %>%
  mutate(results = map(
    .x = iteration, 
    .f = ~ {
    generate_data(n_per_condition = 50,
                  mean_control = 0,
                  mean_intervention = 0.5,
                  sd = 1) %>%
      analyse_data()
  })) %>%
  unnest(results)

head(simulation_results_null_twenty_per_group)
head(simulation_results_null_fifty_per_group)
head(simulation_results_alternative_twenty_per_group)
head(simulation_results_alternative_fifty_per_group)
```

As you can see, this involves copy-pasting, which (as we discussed previously) is something we want to avoid wherever possible.

Our issue is that we have two different parameters (n_per_condition, mean_intervention) that we want to change. To achieve this, we cannot use `map()`, because `map()` can only handle one parameter changing at a time (i.e., `.x`). Instead, we need a `map()` function that *can handle two inputs at once*. Fortunately, `map2()` was made for exactly this purpose! There also exist similar extensions for `map2_*()` as there are for `map_*()`, for example `map2_dfr()` and `map2_dbl()`: these behave identically to their `map_*()` equivalents, but just take two inputs instead of one.

### Use `map2()` to map multple parameters at once

Let's first define our experiment parameters in a tibble. We can do this using `expand_grid()`, as we learned in the previous chapter.

```{r}
set.seed(42) 

n_iterations <- 100 # lowered for testing; 2000 would be more appropriate

experiment_parameters <- expand_grid(
  iteration = 1:n_iterations,
  n_per_condition = c(20, 50),
  mean_intervention = c(0, 0.5),
  mean_control = 0,
  sd = 1
)

experiment_parameters |>
  head(n = 30)
```

We can double-check our output to make sure we're covering all parameter combinations we care about. We can do this by calling `distinct()` on only the columns that we want to vary (and ignore iteration, since this just repeats settings). We can specify `.keep_all = TRUE` in `distinct()` to keep all columns for our inspection; by default, `distinct()` will drop columns that are not mentioned in the `distinct()` function:

```{r}
experiment_parameters %>%
  distinct(n_per_condition, mean_intervention, .keep_all = TRUE)
```

We now have our range of experiments defined: 2000 iterations for each combination, with each condition represented (as shown in the double-check above).

We’ll now run all of this over `map2()`. To do this, we need to specify one input in `map2()` with the .x argument (just like `map()`), and another input with the .y argument. We specify the function (`generate_data()`, in this case) with .f. And somewhat confusingly, all of this becomes the `.x` input for a call to `map()`, with `.f` being our `analyse_data()` function!

```{r}
simulation_results <- experiment_parameters %>%
  mutate(
    results = map(
      .x = map2(
        .x = n_per_condition,
        .y = mean_intervention,
        .f = ~ generate_data(
          n_per_condition   = .x,
          mean_control      = 0,
          mean_intervention = .y,
          sd                = 1
        )
      ), 
      .f = analyse_data)
  ) %>%
  unnest(results)

simulation_results |>
  head(n = 30)
```

We can now look at the distribution of p-values across sample sizes and values for `mean_intervention`, for example in a simple histogram:

```{r}
simulation_results %>%
  ggplot() +
  aes(x = p) +
  geom_histogram() +
  facet_wrap(~mean_intervention + n_per_condition)
```

## Keeping the data 

One of the most important things to consider when writing simulations is to make them as **reproducible** as possible. One way to do this, as we've discussed previously, is to use the `set.seed()` function to ensure that pseudorandomisation occurs in the same manner when another person tries to run our code. But there is one additional thing we can do that can help with reproducibility: storing the data from each simulation run. 

In our example above where we looked at the distribution of p-values, for example, our output tibble consists of the following columns: iteration, n_per_condition, mean_intervention, mean_control, sd, and p. The first 5 columns are input parameters, and the last column (`p`) is our output *after* the `analyse_data()` function is run on the data. But we don't have the data itself: i.e., the tibble created from the `generate_data()` function. Because of this, we cannot clearly trace each step of the simulation process, and this is not ideal for reproducibility.

The solution, fortunately, is simple: we just modify our workflow to store the generated data in the final simulation data. We do this by adding a `mutate()` to add the data created for each iteration to our parameters tibble, and *only then* do we feed this into the `analyse_data()` function. This is one of the many reasons why making sure the output of `generate_data()` is a tibble is useful: the contents of a particular row in a tibble can also be tibble, meaning we can store a full set of generated data in a single cell!

Let's look at how to do this:

```{r}
simulation_results <- experiment_parameters %>%
  mutate(
    generated_data = map2(
        .x = n_per_condition,
        .y = mean_intervention,
        .f = ~ generate_data(
          n_per_condition   = .x,
          mean_control      = 0,
          mean_intervention = .y,
          sd                = 1
        )
      ),
    results = map(
      .x = generated_data, 
      .f = analyse_data)
  ) %>%
  unnest(results)

simulation_results |>
  head(n = 30)
```


Our output is identical to previously, except we now have the additional column `generated_data`, each of which contains a tibble. Let's look at what happens if we run `unnest()` on the `generated_data` column in the first row of the `simulation_results` tibble:

```{r}
simulation_results %>%
  # take the first row only
  slice_head(n = 1) %>%  
  # unnest the generated data column into the tibble that contains it
  unnest(generated_data)
```

We can now inspect the full set of generated data that was simulated for any given row in the `simulation_results`!

We can also access the contents of individual data sets using base-R. Doing it this way can help us understand the nature of nested tibbles; that each cell in the tibble's data column contains another tibble. 

```{r}
data_set_1 <- simulation_results$generated_data[[1]]

data_set_1 
```

In principle, we could re-analyze individual datasets, run other diagnostic tests or checks on them, etc.

```{r}
analyse_data(data_set_1)
```

## Mapping with multiple inputs: `pmap()`

Above, we used `map2()` to map two inputs into a simulation. But there will be many cases where we want to map many different inputs at once - not just one or two.

For example: maybe we want to vary all values of n_per_condition, mean_control, mean_intervention, and sd. We can use the combination of `expand_grid()` and `pmap()` to achieve this.

With `pmap()`, we supply all of the parameters as a list to the .l argument, and our function (`generate_data()`) to the .f argument. We'll use the same approach as we used previously: run `pmap()` within a `mutate()` call and save the output to a `generated_data` column, then `map()` the generated data onto `analyse_data()`, then call `unnest()` on the results. Because we're now dealing with more combinations of parameters, we will lower the value of `n_iterations` from 2000 to 100 - otherwise our code might take a little while to run!

```{r}
n_iterations <- 100 # lowered for testing; 2000 would be more appropriate

# specify the parameters
experiment_parameters <- expand_grid(
  iteration = 1:n_iterations, 
  n_per_condition = c(20, 50),
  mean_intervention = c(0, .5),
  mean_control = c(0, .1),
  sd = c(.5, 1)
)

# use pmap() across this
simulation_results <- experiment_parameters %>%
  mutate(generated_data = pmap(.l = list(n_per_condition, mean_control, mean_intervention, sd),
                          .f = generate_data)) %>%
  mutate(results = map(.x = generated_data, 
                       .f = analyse_data)) %>%
  unnest(results)

simulation_results
```

Let's plot our results and see what the distribution of p-values looks like across all of these different parameter combinations:

```{r}
ggplot(simulation_results, aes(x = p)) +
  geom_histogram(bins = 30) +
  facet_wrap(~ n_per_condition, labeller = label_both) +
  theme_minimal()

ggplot(simulation_results, aes(x = p)) +
  geom_histogram(bins = 30) +
  facet_wrap(~ mean_intervention, labeller = label_both) +
  theme_minimal()

ggplot(simulation_results, aes(x = p)) +
  geom_histogram(bins = 30) +
  facet_wrap(~ mean_control, labeller = label_both) +
  theme_minimal()

ggplot(simulation_results, aes(x = p)) +
  geom_histogram(bins = 30) +
  facet_wrap(~ sd, labeller = label_both) +
  theme_minimal()
```

## Parallel mapping with {furrr}

With the `map()` family of functions we have discussed in this chapter, simulations are run *in sequence*. For example, if we are running 100 iterations, this means that iteration 1 is run first; then iteration 2 is run; then iteration 3; etc.

When you’re doing hundreds of thousands, millions, (or billions!) of iterations, running simulations in sequence can take a long time. In such cases, we ideally would run many of the simulations *in parallel* to speed things up (e.g., running the first 5 iterations at once; then the next 5; etc.). This is called parallelisation, and it can dramatically speed up simulations.

The package for running parallelised simulations is called {furrr}. {furrr} has versions of the `map()` functions that mirror those in {purrr}:

-   `future_map()` is the parallelised version of `map()`
-   `future_map_dfr()` is the parallelised version of `map_dfr()`
-   `future_pmap()` is the parallelised version of `pmap()`

Here is the same approach from above, but using future_pmap(). Since we're parallelising, we'll be ambitious and increase `n_iterations` to 10000. 

```{r}
# dependencies
library(furrr)
library(readr)

# set up parallelization for future/furrr
# use availableCores() to define the number of cores, in case the simulation is being
# run on a HPC that might have more physical cores than are available to the session
plan(multisession, workers = parallelly::availableCores())


n_iterations <- 1000 # lowered for testing; 2000+ would be more appropriate

# specify the parameters
experiment_parameters <- expand_grid(
  iteration = 1:n_iterations,
  n_per_condition = c(20, 50),
  mean_intervention = c(0, .5),
  mean_control = c(0, .1),
  sd = c(.5, 1)
)

# results are cached to disk: the simulation is only run the first time, and read back
# in on subsequent renders. delete the .rds file if you change the simulation code.
dir.create("results", showWarnings = FALSE)

if(file.exists("results/simulation_mapping.rds")){

  simulation_parallel <- read_rds("results/simulation_mapping.rds")

} else {

  simulation_parallel <- experiment_parameters %>%
    # shuffle the rows of the grid to load balance the computationally expensive
    # iterations evenly across the workers
    slice_sample(prop = 1) %>%
    mutate(generated_data = future_pmap(.l = list(n_per_condition, mean_control, mean_intervention, sd),
                                        .f = generate_data,
                                        # this gives us a progress bar so we can see how the simulation is progressing
                                        .progress = TRUE,
                                        .options = furrr_options(seed = TRUE))) %>%
    mutate(results = future_map(.x = generated_data,
                                .f = analyse_data,
                                .progress = TRUE,
                                .options = furrr_options(seed = TRUE))) %>%
    unnest(results) %>%
    # undo the shuffle so the rows are back in the order of the experiment grid
    arrange(n_per_condition, mean_intervention, mean_control, sd, iteration) %>%
    # the generated data sets are not needed after they have been analyzed, and keeping
    # them would make the saved file enormous
    select(-generated_data)

  write_rds(x = simulation_parallel, file = "results/simulation_mapping.rds", compress = "gz")

}
```

Once again, let's plot these results:

```{r}
ggplot(simulation_parallel, aes(x = p)) +
  geom_histogram(bins = 30) +
  facet_wrap(~ n_per_condition, labeller = label_both) +
  theme_minimal()

ggplot(simulation_parallel, aes(x = p)) +
  geom_histogram(bins = 30) +
  facet_wrap(~ mean_intervention, labeller = label_both) +
  theme_minimal()

ggplot(simulation_parallel, aes(x = p)) +
  geom_histogram(bins = 30) +
  facet_wrap(~ mean_control, labeller = label_both) +
  theme_minimal()

ggplot(simulation_parallel, aes(x = p)) +
  geom_histogram(bins = 30) +
  facet_wrap(~ sd, labeller = label_both) +
  theme_minimal()
```

As you might be able to tell already, plotting lots of results like this is hard to read, particularly as the number of parameters manipulated increases. That is why our next chapter will deal with the final step of Monte Carlo simulation: *summarising over the results*.

## Summary

You now have learned one of the core things that makes a simulation a simulation: doing the process of generating data and analysing it *many times*. Specifically, we covered how:

-   `map()` / `map2()` / `pmap()` can be used to repeat processes over different parameters created from `expand_grid()`
-   `unnest()` flattens results for summarizing and plotting
-   `map_dbl()` and `map_dfr()` help you get outputs in the shape you want
-   `future_*()` variants can make big simulations faster

As a take-home message, remember: `pmap()` will be our go-to function for the remainder of this book when it comes to simulation workflows.

::: {.callout-tip title = "Practice writing analysis functions"}
Ready to practice? 
Download and complete the [exercises for this chapter](../exercises/08_mapping_exercises.qmd).
:::
