# Estimation

```{r}
#| include: false
# if it is available, run the setup script that tells quarto to round all df/tibble outputs to three decimal places
if(file.exists("../_setup.R")){source("../_setup.R")}
```

<!-- TODO: Add precision vs bias (accuracy) plots and explanations, with german terms -->
<!-- TODO: Add link to definition of CI like https://www.the100.ci/2024/12/05/why-you-are-not-allowed-to-say-that-your-95-confidence-interval-contains-the-true-parameter-with-a-probability-of-95/ -->
<!-- TODO: Check understanding of pmap and unnest and nested data. -->
<!-- TODO: Add explainer on nested data -->
<!-- TODO: Explain unnest() by relating it to psych data: one row per participant, one nested data frame of trial level data per row. -->
<!-- TODO: Maybe find German terms for bias and accuracy (if they help) -->
<!-- TODO: add examples of map headaches with argument ordering problems, and why we used named lists -->
<!-- TODO: add example of sth like a p value (only) simulation that uses map_dfr() vs pmap() and the relative pros/cons; the former produces simpler results (avoids unnest) but requires you to know the map functions, the latter allows you to use the same code workflow for everything (pmap with named list). -->

```{r}
#| include: false
# dependencies
library(tidyr)
library(dplyr)
library(readr)
library(furrr)
library(parameters)
library(stringr)
library(forcats)
library(ggplot2)
library(scales)
library(ggstance)
library(patchwork)
library(janitor)
library(effectsize)
library(ggstance)
library(knitr)
library(kableExtra)

# set up parallelization
# use availableCores() to define the number of cores, in case the simulation is being
# run on a HPC that might have more physical cores than are available to the session
plan(multisession, workers = parallelly::availableCores())
```

## Simulation

What does this simulation simulate? Specifically: 

- What study design in the individual studies, and which part of the code should you inspect to understand this?
- What data generating process, and which part of the code should you inspect to understand this?
- What analysis, and which part of the code should you inspect to understand this?
- What experiment is set up in the simulation as a whole, and which part of the code should you inspect to understand this? 

The last two questions require us to first gain an understanding of how results are summarized across iterations.

- What is the estimand quantified in the summary of the simulations' results, and which part of the code should you inspect to understand this?
- Putting all the above together, how would you describe the research question asked by the simulation as a whole?

```{r}
# functions for simulation
generate_data <- function(n_per_condition,
                          mean_control,
                          mean_intervention,
                          sd) {
  
  data_control <- 
    tibble(condition = "control",
           score = rnorm(n = n_per_condition, mean = mean_control, sd = sd))
  
  data_intervention <- 
    tibble(condition = "intervention",
           score = rnorm(n = n_per_condition, mean = mean_intervention, sd = sd))
  
  data_combined <- 
    bind_rows(data_control,
              data_intervention) |>
    mutate(condition = factor(condition, level = c("intervention", "control")))
  
  return(data_combined)
}

analyse <- function(data){
  fit <- t.test(
    formula = score ~ condition,
    var.equal = TRUE, # students t test: assumes equal variances
    data = data
  )
  
  results <- fit %>%
    model_parameters() %>%
    as_tibble() %>% 
    janitor::clean_names() %>% 
    # create estimand
    mutate(confidence_interval_width = ci_high - ci_low) |>
    # select columns of interest
    select(mean_diff = difference, # for parameter recovery
           ci_low, # plotting
           ci_high, # plotting
           confidence_interval_width) # estimand
  
  return(results)
}

# simulation parameters
experiment_parameters <- expand_grid(
  n_per_condition = seq(from = 10, to = 100, by = 10),
  mean_control = 4,
  mean_intervention = 5,
  sd = 1.5,
  iteration = 1:1000
) 

# run simulation
# results are cached to disk: the simulation is only run the first time, and read back
# in on subsequent renders. delete the .rds file if you change the simulation code.
dir.create("results", showWarnings = FALSE)

if(file.exists("results/simulation_estimation.rds")){

  simulation <- read_rds("results/simulation_estimation.rds")

} else {

  # set seed
  set.seed(43)

  simulation <- experiment_parameters |>
    # shuffle the rows of the grid to load balance the computationally expensive
    # iterations evenly across the workers
    slice_sample(prop = 1) |>
    mutate(generated_data = future_pmap(.l = list(n_per_condition = n_per_condition,
                                                  mean_control = mean_control,
                                                  mean_intervention = mean_intervention,
                                                  sd = sd),
                                        .f = generate_data,
                                        .progress = TRUE,
                                        .options = furrr_options(seed = TRUE))) |>
    mutate(results = future_pmap(.l = list(data = generated_data),
                                 .f = analyse,
                                 .progress = TRUE,
                                 .options = furrr_options(seed = TRUE))) |>
    # undo the shuffle so the rows are back in the order of the experiment grid
    arrange(n_per_condition, iteration) |>
    # the generated data sets are not needed after they have been analyzed, and
    # keeping them would make the saved file enormous
    select(-generated_data)

  write_rds(x = simulation, file = "results/simulation_estimation.rds", compress = "gz")

}
```

## Iteration-level results

### Just a few iterations

Difference-in-means

```{r fig.height=8, fig.width=6}
simulation |>
  unnest(results) |>
  group_by(n_per_condition) |>
  slice(1:10) |>
  ungroup() |>
  # plot
  ggplot(aes(mean_diff, iteration)) +
  geom_vline(xintercept = 0, linetype = "dotted") +
  geom_vline(xintercept = 1, linetype = "dashed", color = "purple") +
  geom_point() +
  scale_y_continuous(trans = "reverse",
                     breaks = breaks_pretty(n = 10),
                     name = "Study (iteration)") +
  scale_x_continuous(breaks = breaks_pretty(n = 3),
                     #limits = c(0,1),
                     name = "Sample difference-in-means") +
  theme_linedraw() +
  theme(panel.grid.minor.y = element_blank()) +
  ggtitle("Population Mean difference = 1") +
  facet_wrap(~ n_per_condition, ncol = 2)
```

Difference-in-means and their 95% Confidence Intervals

```{r fig.height=8, fig.width=6}
simulation |>
  unnest(results) |>
  group_by(n_per_condition) |>
  slice(1:10) |>
  ungroup() |>
  # plot
  ggplot(aes(mean_diff, iteration, color = ci_low > 0)) +
  geom_vline(xintercept = 0, linetype = "dotted") +
  geom_vline(xintercept = 1, linetype = "dashed", color = "purple") +
  geom_linerangeh(aes(xmin = ci_low, xmax = ci_high)) +
  geom_point() +
  scale_y_continuous(trans = "reverse",
                     breaks = breaks_pretty(n = 10),
                     name = "Study (iteration)") +
  scale_x_continuous(breaks = breaks_pretty(n = 3),
                     #limits = c(0,1),
                     name = "Sample difference-in-means") +
  theme_linedraw() +
  labs(color = "Statistically significant") +
  theme(panel.grid.minor.y = element_blank()) +
  ggtitle("Population Mean difference = 1") +
  facet_wrap(~ n_per_condition, ncol = 2)
```

### All the iterations

Difference-in-means

```{r}
simulation |>
  unnest(results) |>
  # plot
  ggplot(aes(n_per_condition*2, mean_diff)) +
  geom_hline(yintercept = 0, linetype = "dotted") +
  geom_jitter(alpha = 0.25) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "purple") +
  scale_x_continuous(breaks = breaks_pretty(n = 10),
                     name = "Total sample size per study") +
  scale_y_continuous(breaks = breaks_pretty(n = 10),
                     #limits = c(0,1),
                     name = "Sample difference-in-means") +
  theme_linedraw() +
  scale_color_viridis_d(begin = 0.4, end = 0.6) +
  ggtitle("Population Mean difference = 1") 
```

Difference-in-means' 95% Confidence Interval width

```{r}
simulation |>
  unnest(results) |>
  # plot
  ggplot(aes(n_per_condition*2, confidence_interval_width)) +
  geom_jitter(alpha = 0.25) +
  geom_hline(yintercept = 0, linetype = "dotted") +
  scale_x_continuous(breaks = breaks_pretty(n = 10),
                     name = "Total sample size per study") +
  scale_y_continuous(breaks = breaks_pretty(n = 10),
                     #limits = c(0,1),
                     name = "Sample 95% CI width") +
  theme_linedraw() +
  scale_color_viridis_d(begin = 0.3, end = 0.7) +
  ggtitle("Population Mean difference = 1")
```

## Summarise across iterations

Remember: almost without exception, you must `group_by()` all factors that you experimentally manipualated in the `expand_grid()`, otherwise you are combining results across conditions, making it difficult to know what to conclude (i.e., apples-and-oranges comparisons).

```{r}
# summarise results
results_summary <- simulation |>
  unnest(results) |>
  group_by(n_per_condition) |>
  summarize(mean_difference_in_means = mean(mean_diff),
            mean_ci_low = mean(ci_low),
            mean_ci_high = mean(ci_high),
            mean_confidence_interval_width = mean(confidence_interval_width))

# table
# remember: only round results as you're about to print a table
results_summary |>
  mutate_if(is.numeric, round, digits = 1)
```

Difference-in-means

```{r}
ggplot(results_summary, aes(n_per_condition*2, mean_difference_in_means)) +
  geom_point() +
  geom_linerange(aes(ymin = mean_ci_low, ymax = mean_ci_high)) +
  geom_hline(yintercept = 0, linetype = "dotted") +
  scale_x_continuous(breaks = breaks_pretty(n = 10),
                     name = "Total sample size per study") +
  scale_y_continuous(breaks = breaks_pretty(n = 5),
                     #limits = c(0,1),
                     name = "Mean sample difference-in-means\n(and mean 95% CIs)") +
  theme_linedraw() +
  ggtitle("Population difference-in-means = 1")
```

Difference-in-means' 95% Confidence Interval width

```{r}
ggplot(results_summary, aes(n_per_condition*2, mean_confidence_interval_width)) +
  geom_hline(yintercept = 0, linetype = "dotted") +
  #geom_linerange(aes(ymin = min_ci_width, ymax = max_ci_width)) +
  geom_point() +
  #scale_y_continuous(name = "95% CI width\n(mean [min, max])", limits = c(0, NA)) +
  scale_y_continuous(name = "Mean sample 95% CI width", 
                     limits = c(0, NA)) +
  scale_x_continuous(breaks = breaks_pretty(n = 9), 
                     name = "Total sample size per study") +
  theme_linedraw() +
  scale_color_viridis_d(begin = 0.3, end = 0.7) +
  ggtitle("Population Mean difference = 1")
```


