# p-hacking via different forms of selecting reporting

```{r}
#| include: false
# if it is available, run the setup script that tells quarto to round all df/tibble outputs to three decimal places
if(file.exists("../_setup.R")){source("../_setup.R")}
```

## Overview

*p*-hacking is increasing the false-positive rate through analytic choices. These choices are often a) flexible, in that many different options might be tried until significance is found (Gelman called this the "Garden of Forking Paths"), and b) undisclosed. However, it need not be either in a given instance: perhaps the very first analytic strategy tried produces the significant result (but if it hadn't, other strategies would have been tried), or perhaps the author fully discloses the analytic choices but does not make clear to the reader how this undermines the severity of the test (i.e., it takes a very close reading to understand that the results present weak evidence, or the verbal conclusions oversell this).

p-hacking can occur *because of the uniform distribution of p-values under the null hypothesis*. Because all values are equally likely, you just have to keep rolling the dice to eventually find *p* \< .05.

This lesson illustrates a few different examples of one broad form of p-hacking called selective reporting.

```{r}
#| include: false
library(tidyr)
library(dplyr)
library(readr)
library(forcats)
library(stringr)
library(ggplot2)
library(scales)
library(patchwork)
library(tibble)
library(purrr)
library(furrr)
library(simhelpers)
library(faux)
library(janitor)
#library(afex)

# set up parallelization
# use availableCores() to define the number of cores, in case the simulation is being
# run on a HPC that might have more physical cores than are available to the session
plan(multisession, workers = parallelly::availableCores())
```

## Selective reporting of studies

If I run two experiments, and only report the one that "works", I will increase the false positive rate across studies. Perhaps it is because one really is better designed etc than the other. Or perhaps its just a false positive.

What would you guess the false positive rate is for two studies?

Let's simulate it.

### Run simulation

```{r}
generate_data <- function(n_per_condition,
                          mean_control,
                          mean_intervention,
                          sd) {
  
  data_control <- 
    tibble(condition = "control",
           score = rnorm(n = n_per_condition, mean = mean_control, sd = sd))
  
  data_intervention <- 
    tibble(condition = "intervention",
           score = rnorm(n = n_per_condition, mean = mean_intervention, sd = sd))
  
  data_combined <- bind_rows(data_control,
                             data_intervention)
  
  return(data_combined)
}


# define data analysis function ----
analyze <- function(data) {
  
  students_ttest <- t.test(formula = score ~ condition,
                           data = data,
                           var.equal = TRUE,
                           alternative = "two.sided")
  
  res <- tibble(p = students_ttest$p.value)
  
  return(res)
}


# define experiment parameters ----
experiment_parameters <- expand_grid(
  n_per_condition = 100, 
  mean_control = 0,
  mean_intervention = 0,
  sd = 1,
  iteration = 1:1000
)


# run simulation ----
# results are cached to disk: the simulation is only run the first time, and read back
# in on subsequent renders. delete the .rds file if you change the simulation code.
dir.create("results", showWarnings = FALSE)

if(file.exists("results/simulation_p_hacking_two_studies.rds")){

  simulation <- read_rds("results/simulation_p_hacking_two_studies.rds")

} else {

  set.seed(42)

  simulation <-
    # using the experiment parameters
    experiment_parameters |>

    # shuffle the rows of the grid to load balance the computationally expensive
    # iterations evenly across the workers
    slice_sample(prop = 1) |>

    # generate data using the data generating function and the parameters relevant to data generation
    mutate(generated_data_study1 = future_pmap(.l = list(n_per_condition,
                                                         mean_control,
                                                         mean_intervention,
                                                         sd),
                                               .f = generate_data,
                                               .progress = TRUE,
                                               .options = furrr_options(seed = TRUE))) |>
    mutate(generated_data_study2 = future_pmap(.l = list(n_per_condition,
                                                         mean_control,
                                                         mean_intervention,
                                                         sd),
                                               .f = generate_data,
                                               .progress = TRUE,
                                               .options = furrr_options(seed = TRUE))) |>

    # apply the analysis function to the generated data using the parameters relevant to analysis
    mutate(results_study1 = future_pmap(.l = list(data = generated_data_study1),
                                        .f = analyze,
                                        .progress = TRUE,
                                        .options = furrr_options(seed = TRUE))) |>
    mutate(results_study2 = future_pmap(.l = list(data = generated_data_study2),
                                        .f = analyze,
                                        .progress = TRUE,
                                        .options = furrr_options(seed = TRUE))) |>

    # undo the shuffle so the rows are back in the order of the experiment grid
    arrange(iteration) |>

    # the generated data sets are not needed after they have been analyzed, and keeping
    # them would make the saved file enormous
    select(-generated_data_study1, -generated_data_study2)

  write_rds(x = simulation,
            file = "results/simulation_p_hacking_two_studies.rds",
            compress = "gz")

}


# summarise simulation results over the iterations ----
simulation_summary <- simulation |>
  # unnest and rename
  unnest(results_study1) |>
  rename(p_study1 = p) |>
  unnest(results_study2) |>
  rename(p_study2 = p) |>
  # simulate flexible reporting
  mutate(p_hacked = ifelse(p_study1 < .05, p_study1, p_study2)) |>
  # summarize. rather than calculating the proportion of significant results by hand,
  # use {simhelpers}'s dedicated function for the empirical rejection rate.
  # calc_rejection() also returns the number of iterations and the Monte Carlo Standard
  # Error of the rate; select() keeps and renames just the rate itself
  reframe(calc_rejection(data     = pick(everything()),
                         p_values = p_study1,
                         alpha    = 0.05) |>
            select(prop_sig_study1 = rej_rate),
          calc_rejection(data     = pick(everything()),
                         p_values = p_study2,
                         alpha    = 0.05) |>
            select(prop_sig_study2 = rej_rate),
          calc_rejection(data     = pick(everything()),
                         p_values = p_hacked,
                         alpha    = 0.05) |>
            select(prop_sig_hacked = rej_rate)) |>
  pivot_longer(cols = everything(),
               names_to = "Source",
               values_to = "proportion_significant") |>
  mutate(Source = str_remove(Source, "prop_sig_"))
```

### Results

```{r}
simulation_summary |>
  mutate_if(is.numeric, round_half_up, digits = 2) 
```

### Conclusions

The false positive rate for two studies is \~10%, because each study has a 5% false positive rate.

See the exercises for this chapter to extend this simulation to four studies.

## Familywise error in uncorrelated tests

Note that these probabilities are merely additive. This might not be intuitive.

While the false positive rate for any individual test is 5%, the probability of finding at least one significant results in exactly $k$ number of tests follows the equation:

$$
FPR = 1-(1-\alpha)^k,
$$

where $\alpha$ is the alpha value for the tests (e.g., 0.05) and $k$ is the number of tests run.

We can write a function to calculate the FPR mathematically for different values of $\alpha$ and $k$.

```{r}
fpr <- function(k, alpha = 0.05){
  fpr <- 1 - (1 - alpha)^k
  return(fpr)
}

dat_fpr <- tibble(k = seq(from = 1, to = 10, by = 1)) |>
  mutate(fpr = map_dbl(k, fpr)) 

dat_fpr |>
  mutate_if(is.numeric, round_half_up, digits = 2)
```

This problem of familywise error is common in ANOVA, where there are main and interaction effects (see Cramer, A.O.J., van Ravenzwaaij, D., Matzke, D. et al. Hidden multiplicity in exploratory multiway ANOVA: Prevalence and remedies. Psychon Bull Rev 23, 640–647 (2016). https://doi.org/10.3758/s13423-015-0913-5).

E.g., 2X2 ANOVA:

- Main effect for condition 1
- Main effect for condition 2
- Interaction effect

2X2X2 ANOVA?

2X2X2X2 ANOVA?

```{r}

n_pvalues <- function(n_factors){
  2^n_factors - 1
}

n_pvalues(3)

n_pvalues(4)

```


```{r}

n_pvalues(3) |>
  fpr()

n_pvalues(4) |>
  fpr()

```

What about situations where the outcomes are not uncorrelated, such as anxiety, panic attacks, and other psychological symptoms? E.g., Gloster et al. (2015, doi [10.1007/s00406-015-0575-3](https://doi.org/10.1007/s00406-015-0575-3)) which examines differences in means between participants with and without the 5HTT-LPR polymorphism on 5 likely correlated outcome variables?

The reported results aren't robust to reasonable corrections for multiple testing:

```{r}
ps <- c(0.013, 0.460, 0.924, 0.864, 0.650)

p.adjust(p = ps, method = "holm")
```

This is unsurprising, since a very large scale preregistered study shows no associations with 5HTT-LPR (Border et al., 2019, doi [10.1176/appi.ajp.2018.18070881](https://doi.org/10.1176/appi.ajp.2018.18070881)) and that research area has been substantively discredited (as has the candidate gene literature in general).

But much do these five correlated outcome measures affect the false positive rate, specifically?

We can solve this with math, but more complex things require either much more complicated math or ... simulation.

## Selective reporting of correlated outcomes

Aka "outcome switching", which is both common and problematic in clinical trials.

There are of course situations where the tests are not independent, and this makes it more complicated to understand the FPR. For example, imagine I am interested in the impact of CBT on depression, but I include multiple measures of depression and report the one that 'worked'.

### Generate correlated data

We haven't simulated correlated data before so lets do that first.

```{r fig.height=5, fig.width=5}

set.seed(42)

dat <- 
  faux::rnorm_multi(n = 1000,
                    mu = c(0, 0),
                    sd = c(1, 1),
                    r = -0.7,
                    varnames = c("var1", "var2"))

cor(dat)

ggplot(dat, aes(var1, var2)) +
  geom_point(alpha = 0.4) +
  theme_linedraw()

```

### Generate data

Correlated outcomes, known groups differences.

```{r}

generate_data <- function(n_per_condition,
                          r_between_outcomes,
                          mu_control, # vector of length 1 or k
                          mu_intervention, # vector of length 1 or k
                          sigma_control = 1, # vector of length 1 or k
                          sigma_intervention = 1) {  # vector of length 1 or k
  
  # generate data by condition
  data_control <- 
    rnorm_multi(n = n_per_condition,
                mu = mu_control,
                sd = sigma_control,
                r = r_between_outcomes,
                varnames = c("outcome1", "outcome2", "outcome3")) |>
    mutate(condition = "Control")
  
  data_intervention <- 
    rnorm_multi(n = n_per_condition,
                mu = mu_intervention,
                sd = sigma_intervention,
                r = r_between_outcomes,
                varnames = c("outcome1", "outcome2", "outcome3")) |>
    mutate(condition = "Intervention")
  
  # combine
  data_combined <- 
    bind_rows(data_control, 
              data_intervention) |>
    mutate(condition = fct_relevel(condition, "Intervention", "Control"))
  
  return(data_combined)  
}

```

Check the function works

```{r}

# simulate data
set.seed(42)

dat <- generate_data(n_per_condition = 100000,
                     r_between_outcomes = 0.6,
                     mu_control = 0,
                     mu_intervention = 1.5)

# parameter recovery
dat |>
  group_by(condition) |>
  summarize(mean_outcome1 = mean(outcome1),
            mean_outcome2 = mean(outcome2),
            correlation = cor(outcome1, outcome2)) |>
  mutate_if(is.numeric, round_half_up, digits = 2)

```

### Run simulation

```{r}

# define data analysis function ----
analyze <- function(data, outcome) {
  
  # if you want to specify a column as an input, use rename()+{{}} to rename it before analysis
  data_renamed <- data |>
    rename(score = {{ outcome }}) 
  
  students_ttest <- t.test(formula = score ~ condition,
                           data = data_renamed,
                           var.equal = TRUE,
                           alternative = "two.sided")
  
  res <- tibble(p = students_ttest$p.value)
  
  return(res)
}


# define experiment parameters ----
experiment_parameters <- expand_grid(
  n_per_condition = 50, 
  r_between_outcomes = c(0, .1, .2, .3, .4, .5, .6, .7, .8, .9, .99),
  mu_control = 0,
  mu_intervention = 0,
  iteration = 1:20000
) 


# run simulation ----
# results are cached to disk: the simulation is only run the first time, and read back
# in on subsequent renders. delete the .rds file if you change the simulation code.
dir.create("results", showWarnings = FALSE)

if(file.exists("results/simulation_p_hacking_correlated_outcomes.rds")){

  simulation <- read_rds("results/simulation_p_hacking_correlated_outcomes.rds")

} else {

  set.seed(42)

  simulation <-
    # using the experiment parameters
    experiment_parameters |>

    # shuffle the rows of the grid to load balance the computationally expensive
    # iterations evenly across the workers
    slice_sample(prop = 1) |>

    # generate data using the data generating function and the parameters relevant to data generation
    mutate(generated_data = future_pmap(.l = list(n_per_condition,
                                                  r_between_outcomes,
                                                  mu_control,
                                                  mu_intervention),
                                        .f = generate_data,
                                        .progress = TRUE,
                                        .options = furrr_options(seed = TRUE))) |>

    # apply the analysis function to the generated data using the parameters relevant to analysis
    mutate(results_outcome1 = future_pmap(.l = list(data = generated_data,
                                                    outcome = "outcome1"),
                                          .f = analyze,
                                          .progress = TRUE,
                                          .options = furrr_options(seed = TRUE)),
           results_outcome2 = future_pmap(.l = list(data = generated_data,
                                                    outcome = "outcome2"),
                                          .f = analyze,
                                          .progress = TRUE,
                                          .options = furrr_options(seed = TRUE)),
           results_outcome3 = future_pmap(.l = list(data = generated_data,
                                                    outcome = "outcome3"),
                                          .f = analyze,
                                          .progress = TRUE,
                                          .options = furrr_options(seed = TRUE))) |>

    # unnest and rename
    unnest(results_outcome1) |>
    rename(outcome1_p = p) |>
    unnest(results_outcome2) |>
    rename(outcome2_p = p) |>
    unnest(results_outcome3) |>
    rename(outcome3_p = p) |>

    # undo the shuffle so the rows are back in the order of the experiment grid
    arrange(r_between_outcomes, iteration) |>

    # the generated data sets are not needed after they have been analyzed, and keeping
    # them would make the saved file enormous
    select(-generated_data)

  write_rds(x = simulation,
            file = "results/simulation_p_hacking_correlated_outcomes.rds",
            compress = "gz")

}


# summarise simulation results over the iterations ----
simulation_summary <- simulation |>
  # the smallest of the three p values: it is below alpha exactly when at least one of
  # the three outcomes was significant, i.e. it is the p value a researcher who switches
  # outcomes would report
  mutate(p_smallest = pmin(outcome1_p, outcome2_p, outcome3_p)) |>
  # summarize across iterations: empirical detection rates, using {simhelpers}'s
  # dedicated function for the empirical rejection rate
  group_by(n_per_condition,
           r_between_outcomes,
           mu_control,
           mu_intervention) |>
  reframe(calc_rejection(data = pick(everything()), p_values = outcome1_p, alpha = 0.05) |>
            select(prop_sig_outcome1 = rej_rate),
          calc_rejection(data = pick(everything()), p_values = outcome2_p, alpha = 0.05) |>
            select(prop_sig_outcome2 = rej_rate),
          calc_rejection(data = pick(everything()), p_values = outcome3_p, alpha = 0.05) |>
            select(prop_sig_outcome3 = rej_rate),
          calc_rejection(data = pick(everything()), p_values = p_smallest, alpha = 0.05) |>
            select(prop_sig_at_least_one = rej_rate))

```

### Results

Empirical false positive rates as a function of the true correlation between outcomes. `prop_sig_at_least_one` is the empirical detection rate — the proportion of iterations in which *at least one* of the three outcomes reached *p* \< .05, i.e. what a researcher who switches outcomes would report as "significant".

```{r}

simulation_summary |>
  mutate_if(is.numeric, round_half_up, digits = 3)

ggplot(simulation_summary, aes(r_between_outcomes, prop_sig_outcome1)) +
  geom_hline(yintercept = 0.05, linetype = "dotted") +
  geom_point() +
  geom_line() +
  scale_x_continuous(breaks = breaks_pretty(n = 9), name = "True correlation between outcomes") +
  scale_y_continuous(limits = c(0, .2), breaks = breaks_pretty(n = 5), name = "Empirical Detection Rate\n(outcome 1)") +
  theme_linedraw() +
  scale_color_viridis_d(begin = 0.2, end = 0.8)

ggplot(simulation_summary, aes(r_between_outcomes, prop_sig_outcome2)) +
  geom_hline(yintercept = 0.05, linetype = "dotted") +
  geom_point() +
  geom_line() +
  scale_x_continuous(breaks = breaks_pretty(n = 9), name = "True correlation between outcomes") +
  scale_y_continuous(limits = c(0, .2), breaks = breaks_pretty(n = 5), name = "Empirical Detection Rate\n(outcome 2)") +
  theme_linedraw() +
  scale_color_viridis_d(begin = 0.2, end = 0.8)

ggplot(simulation_summary, aes(r_between_outcomes, prop_sig_outcome3)) +
  geom_hline(yintercept = 0.05, linetype = "dotted") +
  geom_point() +
  geom_line() +
  scale_x_continuous(breaks = breaks_pretty(n = 9), name = "True correlation between outcomes") +
  scale_y_continuous(limits = c(0, .2), breaks = breaks_pretty(n = 5), name = "Empirical Detection Rate\n(outcome 3)") +
  theme_linedraw() +
  scale_color_viridis_d(begin = 0.2, end = 0.8)

ggplot(simulation_summary, aes(r_between_outcomes, prop_sig_at_least_one)) +
  geom_hline(yintercept = 0.05, linetype = "dotted") +
  geom_point() +
  geom_line() +
  scale_x_continuous(breaks = breaks_pretty(n = 9), name = "True correlation between outcomes") +
  scale_y_continuous(limits = c(0, .2), breaks = breaks_pretty(n = 5), name = "Empirical Detection Rate\n(at least one outcome significant)") +
  theme_linedraw() +
  scale_color_viridis_d(begin = 0.2, end = 0.8)

```

### Conclusions

How does the false positive rate change when the outcome measures are correlated?

What perverse incentive does this set up for researchers?

## Check your learning

If authors reported the results of all outcome variables (and not just the ones that 'worked'):

- Would this problem be improved?
- Would this problem be completely solved? 

See the exercises for this chapter for further practice.

## Readings

-   Stefan, A. M. and Schönbrodt F. D. (2023) Big little lies: a compendium and simulation of p-hacking strategies. Royal Society Open Science. 10220346 http://doi.org/10.1098/rsos.220346
