# The differences between significant and non-significant is not itself significant <span class="badge badge-draft1">✎ Very rough draft</span>

```{r}
#| include: false
# if it is available, run the setup script that tells quarto to round all df/tibble outputs to three decimal places
if(file.exists("../_setup.R")){source("../_setup.R")}
```

## Dependencies

```{r}

library(tidyr)
library(dplyr)
library(readr)
library(purrr)
library(furrr)
library(simhelpers)
library(stringr)
library(forcats)
library(ggplot2)
library(scales)
library(patchwork)
library(janitor)
library(effsize)
library(metafor)

# set up parallelization
# use availableCores() to define the number of cores, in case the simulation is being
# run on a HPC that might have more physical cores than are available to the session
plan(multisession, workers = parallelly::availableCores())

format_p_apa <- function(p, threshold = 0.001) {
  ifelse(
    p < threshold,
    paste0("p < ", sub("^0\\.", ".", format(threshold, nsmall = 3))),
    paste0("p = ", sub("^0\\.", ".", format(round(p, 3), nsmall = 3)))
  )
}

```

## Simulation functions

```{r}

# functions for simulation
generate_data <- function(n_per_condition,
                          population_cohens_d) {
  
  data_control <- 
    tibble(condition = "control",
           score = rnorm(n = n_per_condition, mean = 0, sd = 1))
  
  data_intervention <- 
    tibble(condition = "intervention",
           score = rnorm(n = n_per_condition, mean = population_cohens_d, sd = 1))
  
  data_combined <- bind_rows(data_control,
                             data_intervention) |>
    mutate(condition = fct_relevel(condition, "intervention", "control"))
  
  return(data_combined)
}

analyze <- function(data, study) {

  res_t_test <- t.test(formula = score ~ condition, 
                       data = data,
                       var.equal = TRUE,
                       alternative = "two.sided")
  
  res_cohens_d <- effsize::cohen.d(formula = score ~ condition,
                                   data = data,
                                   pooled = TRUE)
  
  res <- tibble(study = study,
                p = res_t_test$p.value, 
                cohens_d = res_cohens_d$estimate,
                cohens_d_ci_lower = res_cohens_d$conf.int[1],
                cohens_d_ci_upper = res_cohens_d$conf.int[2],
                cohens_d_se = sqrt(res_cohens_d$var))

  return(res)
}

meta_analyze <- function(results_study_1, results_study_2) {
  
  dat <- bind_rows(results_study_1,
                   results_study_2)
  
  fit <- rma(data = dat, 
             yi = cohens_d, 
             sei = cohens_d_se, 
             mods = ~ study, 
             method = "FE")

  res <- tibble(cohens_d = fit$b[2],
                cohens_d_ci_lower = fit$ci.lb[2],
                cohens_d_ci_upper = fit$ci.ub[2],
                #df = fit$QMdf[1],
                #Q = fit$QM,
                p = fit$QMp) # p of moderation
    #mutate(formatted = paste0("Q(", df, ") = ", round_half_up(Q, 3), ", ", format_p_apa(p)))

  return(res)
}

```

## Different population effect, large N

```{r fig.height=5, fig.width=14}

# set seed
set.seed(42)

# simulation parameters
experiment_parameters <- expand_grid(
  study_1_n_per_condition = 300,
  study_1_population_cohens_d = 0.5,
  study_2_n_per_condition = 300,
  study_2_population_cohens_d = 0,
  iteration = 1:1000
) 

# run simulation
# results are cached to disk: the simulation is only run the first time, and read back
# in on subsequent renders. delete the .rds file if you change the simulation code.
dir.create("results", showWarnings = FALSE)

if(file.exists("results/simulation_sig_vs_nonsig_1.rds")){

  simulation <- read_rds("results/simulation_sig_vs_nonsig_1.rds")

} else {

  simulation <- experiment_parameters |>
  # shuffle the rows of the grid to load balance the computationally expensive
  # iterations evenly across the workers
  slice_sample(prop = 1) |>
  mutate(generated_data_study_1 = future_pmap(.l = list(n_per_condition = study_1_n_per_condition,
                                                        population_cohens_d = study_1_population_cohens_d),
                                              .f = generate_data,
                                              .progress = TRUE,
                                              .options = furrr_options(seed = TRUE))) |>
  mutate(generated_data_study_2 = future_pmap(.l = list(n_per_condition = study_2_n_per_condition,
                                                        population_cohens_d = study_2_population_cohens_d),
                                              .f = generate_data,
                                              .progress = TRUE,
                                              .options = furrr_options(seed = TRUE))) |>
  mutate(results_study1 = future_pmap(.l = list(data = generated_data_study_1),
                                      .f = analyze,
                                      study = "study 1",
                                      .progress = TRUE,
                                      .options = furrr_options(seed = TRUE))) |>
  mutate(results_study2 = future_pmap(.l = list(data = generated_data_study_2),
                                      .f = analyze,
                                      study = "study 2",
                                      .progress = TRUE,
                                      .options = furrr_options(seed = TRUE))) |>
  mutate(results_meta = future_pmap(.l = list(results_study_1 = results_study1,
                                              results_study_2 = results_study2),
                                    .f = meta_analyze,
                                    .progress = TRUE,
                                    .options = furrr_options(seed = TRUE))) |>
  # undo the shuffle so the rows are back in the order of the experiment grid
  arrange(iteration) |>
  # wrangle results
  unnest(results_study1, names_sep = "_") |>
  unnest(results_study2, names_sep = "_") |>
  unnest(results_meta, names_sep = "_") |>
  select(-generated_data_study_1, -generated_data_study_2) |>
  # this is a complex pivot, don't worry if you don't immediately understand it
  pivot_longer(
    cols = c(
      results_study1_p,
      results_study1_cohens_d,
      results_study1_cohens_d_ci_lower,
      results_study1_cohens_d_ci_upper,
      results_study1_cohens_d_se,
      results_study2_p,
      results_study2_cohens_d,
      results_study2_cohens_d_se,
      results_study2_cohens_d_ci_lower,
      results_study2_cohens_d_ci_upper,
      results_meta_p,
      results_meta_cohens_d,
      # results_meta_cohens_d_se, # not created
      results_meta_cohens_d_ci_lower,
      results_meta_cohens_d_ci_upper
    ),
    names_to = c("source", "metric"),
    names_pattern = "results_(study\\d+|meta)_(.+)",
    values_to = "value"
  ) |>
  pivot_wider(
    names_from = metric,
    values_from = value
  )

  write_rds(x = simulation, file = "results/simulation_sig_vs_nonsig_1.rds", compress = "gz")

}

simulation_summary <- simulation |>
  group_by(study_1_n_per_condition,
           study_1_population_cohens_d,
           study_2_n_per_condition,
           study_2_population_cohens_d,
           source) |>
  # rather than calculating the proportion of significant results by hand, use
  # {simhelpers}'s dedicated function for the empirical rejection rate.
  # calc_rejection() also returns the number of iterations and the Monte Carlo Standard
  # Error of the rate; select() keeps and renames just the rate itself
  reframe(calc_rejection(data     = pick(everything()),
                         p_values = p,
                         alpha    = 0.05) |>
            select(proportion_significant = rej_rate),
          mean_cohens_d = mean(cohens_d),
          mean_cohens_d_ci_lower = mean(cohens_d_ci_lower),
          mean_cohens_d_ci_upper = mean(cohens_d_ci_upper)) |>
  mutate(source = str_replace(source, "meta", "difference")) |>
  mutate(source = fct_relevel(source, "study1", "study2", "difference"))


p_effect_size <- ggplot(simulation_summary, aes(source, mean_cohens_d)) +
  geom_hline(yintercept = 0, linetype = "dotted") +
  geom_linerange(aes(ymin = mean_cohens_d_ci_lower, ymax = mean_cohens_d_ci_upper)) +
  geom_point() +
  scale_y_continuous(breaks = breaks_pretty(n = 10),
                     name = "Mean Cohen's d") +
  xlab("") +
  theme_linedraw() +
  theme(panel.grid.minor = element_blank())

p_proportion_significant <- ggplot(simulation_summary, aes(source, proportion_significant)) +
  geom_hline(yintercept = 0.05, linetype = "dashed") +
  geom_hline(yintercept = 0.80, linetype = "dashed") +
  geom_bar(stat = "identity") +
  scale_y_continuous(breaks = breaks_pretty(n = 10), limits = c(0, 1), name = "Proportion of significant p values") +
  xlab("") +
  # geom_text(aes(x = "study1", y = 1.1, label = paste("Population\nCohen's d =", study_1_population_cohens_d))) +
  # geom_text(aes(x = "study2", y = 1.1, label = paste("Population\nCohen's d =", study_2_population_cohens_d))) +
  # geom_text(aes(x = "difference", y = 1.1, label = paste("Population\nCohen's d =", study_2_population_cohens_d - study_1_population_cohens_d))) +
  theme_linedraw() +
  theme(panel.grid.minor = element_blank())

p_effect_size + p_proportion_significant + plot_layout(ncol = 2)

```

## Different population effect, small N

```{r fig.height=5, fig.width=14}

# set seed
set.seed(42)

# simulation parameters
experiment_parameters <- expand_grid(
  study_1_n_per_condition = 40,
  study_1_population_cohens_d = 0.5,
  study_2_n_per_condition = 40,
  study_2_population_cohens_d = 0,
  iteration = 1:1000
) 

# run simulation
# results are cached to disk: the simulation is only run the first time, and read back
# in on subsequent renders. delete the .rds file if you change the simulation code.
dir.create("results", showWarnings = FALSE)

if(file.exists("results/simulation_sig_vs_nonsig_2.rds")){

  simulation <- read_rds("results/simulation_sig_vs_nonsig_2.rds")

} else {

  simulation <- experiment_parameters |>
  # shuffle the rows of the grid to load balance the computationally expensive
  # iterations evenly across the workers
  slice_sample(prop = 1) |>
  mutate(generated_data_study_1 = future_pmap(.l = list(n_per_condition = study_1_n_per_condition,
                                                        population_cohens_d = study_1_population_cohens_d),
                                              .f = generate_data,
                                              .progress = TRUE,
                                              .options = furrr_options(seed = TRUE))) |>
  mutate(generated_data_study_2 = future_pmap(.l = list(n_per_condition = study_2_n_per_condition,
                                                        population_cohens_d = study_2_population_cohens_d),
                                              .f = generate_data,
                                              .progress = TRUE,
                                              .options = furrr_options(seed = TRUE))) |>
  mutate(results_study1 = future_pmap(.l = list(data = generated_data_study_1),
                                      .f = analyze,
                                      study = "study 1",
                                      .progress = TRUE,
                                      .options = furrr_options(seed = TRUE))) |>
  mutate(results_study2 = future_pmap(.l = list(data = generated_data_study_2),
                                      .f = analyze,
                                      study = "study 2",
                                      .progress = TRUE,
                                      .options = furrr_options(seed = TRUE))) |>
  mutate(results_meta = future_pmap(.l = list(results_study_1 = results_study1,
                                              results_study_2 = results_study2),
                                    .f = meta_analyze,
                                    .progress = TRUE,
                                    .options = furrr_options(seed = TRUE))) |>
  # undo the shuffle so the rows are back in the order of the experiment grid
  arrange(iteration) |>
  # wrangle results
  unnest(results_study1, names_sep = "_") |>
  unnest(results_study2, names_sep = "_") |>
  unnest(results_meta, names_sep = "_") |>
  select(-generated_data_study_1, -generated_data_study_2) |>
  # this is a complex pivot, don't worry if you don't immediately understand it
  pivot_longer(
    cols = c(
      results_study1_p,
      results_study1_cohens_d,
      results_study1_cohens_d_ci_lower,
      results_study1_cohens_d_ci_upper,
      results_study1_cohens_d_se,
      results_study2_p,
      results_study2_cohens_d,
      results_study2_cohens_d_se,
      results_study2_cohens_d_ci_lower,
      results_study2_cohens_d_ci_upper,
      results_meta_p,
      results_meta_cohens_d,
      # results_meta_cohens_d_se, # not created
      results_meta_cohens_d_ci_lower,
      results_meta_cohens_d_ci_upper
    ),
    names_to = c("source", "metric"),
    names_pattern = "results_(study\\d+|meta)_(.+)",
    values_to = "value"
  ) |>
  pivot_wider(
    names_from = metric,
    values_from = value
  )

  write_rds(x = simulation, file = "results/simulation_sig_vs_nonsig_2.rds", compress = "gz")

}

simulation_summary <- simulation |>
  group_by(study_1_n_per_condition,
           study_1_population_cohens_d,
           study_2_n_per_condition,
           study_2_population_cohens_d,
           source) |>
  # rather than calculating the proportion of significant results by hand, use
  # {simhelpers}'s dedicated function for the empirical rejection rate.
  # calc_rejection() also returns the number of iterations and the Monte Carlo Standard
  # Error of the rate; select() keeps and renames just the rate itself
  reframe(calc_rejection(data     = pick(everything()),
                         p_values = p,
                         alpha    = 0.05) |>
            select(proportion_significant = rej_rate),
          mean_cohens_d = mean(cohens_d),
          mean_cohens_d_ci_lower = mean(cohens_d_ci_lower),
          mean_cohens_d_ci_upper = mean(cohens_d_ci_upper)) |>
  mutate(source = str_replace(source, "meta", "difference")) |>
  mutate(source = fct_relevel(source, "study1", "study2", "difference"))


p_effect_size <- ggplot(simulation_summary, aes(source, mean_cohens_d)) +
  geom_hline(yintercept = 0, linetype = "dotted") +
  geom_linerange(aes(ymin = mean_cohens_d_ci_lower, ymax = mean_cohens_d_ci_upper)) +
  geom_point() +
  scale_y_continuous(breaks = breaks_pretty(n = 10),
                     name = "Mean Cohen's d") +
  xlab("") +
  theme_linedraw() +
  theme(panel.grid.minor = element_blank())

p_proportion_significant <- ggplot(simulation_summary, aes(source, proportion_significant)) +
  geom_hline(yintercept = 0.05, linetype = "dashed") +
  geom_hline(yintercept = 0.80, linetype = "dashed") +
  geom_bar(stat = "identity") +
  scale_y_continuous(breaks = breaks_pretty(n = 10), limits = c(0, 1), name = "Proportion of significant p values") +
  xlab("") +
  # geom_text(aes(x = "study1", y = 1.1, label = paste("Population\nCohen's d =", study_1_population_cohens_d))) +
  # geom_text(aes(x = "study2", y = 1.1, label = paste("Population\nCohen's d =", study_2_population_cohens_d))) +
  # geom_text(aes(x = "difference", y = 1.1, label = paste("Population\nCohen's d =", study_2_population_cohens_d - study_1_population_cohens_d))) +
  theme_linedraw() +
  theme(panel.grid.minor = element_blank())

p_effect_size + p_proportion_significant + plot_layout(ncol = 2)

```

## Different population effect, small original study and large "replication"

NB not really a replication as the second study has a different effect size, but its a very rough simulation assuming p hacking took place in the original study and boosted the effect size

\[needs work to be clearer\]

```{r fig.height=10, fig.width=7}

# set seed
set.seed(42)

# simulation parameters
experiment_parameters <- expand_grid(
  study_1_n_per_condition = 40,
  study_1_population_cohens_d = 0.5,
  study_2_n_per_condition = 200,
  study_2_population_cohens_d = 0,
  iteration = 1:1000
) 

# run simulation
# results are cached to disk: the simulation is only run the first time, and read back
# in on subsequent renders. delete the .rds file if you change the simulation code.
dir.create("results", showWarnings = FALSE)

if(file.exists("results/simulation_sig_vs_nonsig_3.rds")){

  simulation <- read_rds("results/simulation_sig_vs_nonsig_3.rds")

} else {

  simulation <- experiment_parameters |>
  # shuffle the rows of the grid to load balance the computationally expensive
  # iterations evenly across the workers
  slice_sample(prop = 1) |>
  mutate(generated_data_study_1 = future_pmap(.l = list(n_per_condition = study_1_n_per_condition,
                                                        population_cohens_d = study_1_population_cohens_d),
                                              .f = generate_data,
                                              .progress = TRUE,
                                              .options = furrr_options(seed = TRUE))) |>
  mutate(generated_data_study_2 = future_pmap(.l = list(n_per_condition = study_2_n_per_condition,
                                                        population_cohens_d = study_2_population_cohens_d),
                                              .f = generate_data,
                                              .progress = TRUE,
                                              .options = furrr_options(seed = TRUE))) |>
  mutate(results_study1 = future_pmap(.l = list(data = generated_data_study_1),
                                      .f = analyze,
                                      study = "study 1",
                                      .progress = TRUE,
                                      .options = furrr_options(seed = TRUE))) |>
  mutate(results_study2 = future_pmap(.l = list(data = generated_data_study_2),
                                      .f = analyze,
                                      study = "study 2",
                                      .progress = TRUE,
                                      .options = furrr_options(seed = TRUE))) |>
  mutate(results_meta = future_pmap(.l = list(results_study_1 = results_study1,
                                              results_study_2 = results_study2),
                                    .f = meta_analyze,
                                    .progress = TRUE,
                                    .options = furrr_options(seed = TRUE))) |>
  # undo the shuffle so the rows are back in the order of the experiment grid
  arrange(iteration) |>
  # wrangle results
  unnest(results_study1, names_sep = "_") |>
  unnest(results_study2, names_sep = "_") |>
  unnest(results_meta, names_sep = "_") |>
  select(-generated_data_study_1, -generated_data_study_2) |>
  # this is a complex pivot, don't worry if you don't immediately understand it
  pivot_longer(
    cols = c(
      results_study1_p,
      results_study1_cohens_d,
      results_study1_cohens_d_ci_lower,
      results_study1_cohens_d_ci_upper,
      results_study1_cohens_d_se,
      results_study2_p,
      results_study2_cohens_d,
      results_study2_cohens_d_se,
      results_study2_cohens_d_ci_lower,
      results_study2_cohens_d_ci_upper,
      results_meta_p,
      results_meta_cohens_d,
      # results_meta_cohens_d_se, # not created
      results_meta_cohens_d_ci_lower,
      results_meta_cohens_d_ci_upper
    ),
    names_to = c("source", "metric"),
    names_pattern = "results_(study\\d+|meta)_(.+)",
    values_to = "value"
  ) |>
  pivot_wider(
    names_from = metric,
    values_from = value
  )

  write_rds(x = simulation, file = "results/simulation_sig_vs_nonsig_3.rds", compress = "gz")

}

simulation_summary <- simulation |>
  group_by(study_1_n_per_condition,
           study_1_population_cohens_d,
           study_2_n_per_condition,
           study_2_population_cohens_d,
           source) |>
  # rather than calculating the proportion of significant results by hand, use
  # {simhelpers}'s dedicated function for the empirical rejection rate.
  # calc_rejection() also returns the number of iterations and the Monte Carlo Standard
  # Error of the rate; select() keeps and renames just the rate itself
  reframe(calc_rejection(data     = pick(everything()),
                         p_values = p,
                         alpha    = 0.05) |>
            select(proportion_significant = rej_rate),
          mean_cohens_d = mean(cohens_d),
          mean_cohens_d_ci_lower = mean(cohens_d_ci_lower),
          mean_cohens_d_ci_upper = mean(cohens_d_ci_upper)) |>
  mutate(source = str_replace(source, "meta", "difference")) |>
  mutate(source = fct_relevel(source, "study1", "study2", "difference"))


p_effect_size <- ggplot(simulation_summary, aes(source, mean_cohens_d)) +
  geom_hline(yintercept = 0, linetype = "dotted") +
  geom_linerange(aes(ymin = mean_cohens_d_ci_lower, ymax = mean_cohens_d_ci_upper)) +
  geom_point() +
  scale_y_continuous(breaks = breaks_pretty(n = 10),
                     name = "Mean Cohen's d") +
  xlab("") +
  theme_linedraw() +
  theme(panel.grid.minor = element_blank())

p_proportion_significant <- ggplot(simulation_summary, aes(source, proportion_significant)) +
  geom_hline(yintercept = 0.05, linetype = "dashed") +
  geom_hline(yintercept = 0.80, linetype = "dashed") +
  geom_bar(stat = "identity") +
  scale_y_continuous(breaks = breaks_pretty(n = 10), limits = c(0, 1), name = "Proportion of significant p values") +
  xlab("") +
  # geom_text(aes(x = "study1", y = 1.1, label = paste("Population\nCohen's d =", study_1_population_cohens_d))) +
  # geom_text(aes(x = "study2", y = 1.1, label = paste("Population\nCohen's d =", study_2_population_cohens_d))) +
  # geom_text(aes(x = "difference", y = 1.1, label = paste("Population\nCohen's d =", study_2_population_cohens_d - study_1_population_cohens_d))) +
  theme_linedraw() +
  theme(panel.grid.minor = element_blank())

p_effect_size + p_proportion_significant + plot_layout(ncol = 1)

```
