# Foundation concepts

## One line Monte Carlo simulation

Answering the question "what is the distribution of *p*-values under the null hypothesis?"

```{r}

set.seed(42)

hist(replicate(100000, t.test(rnorm(n = 50, m = 0, sd = 1), rnorm(n = 50, m = 0, sd = 1))$p.value))

```

This is a complete Monte Carlo simulation. But how is it constructed? Let's built it up more slowly.

## Generate normally distributed data

Draw data from a normally distributed population and plot it using `rnorm()`

```{r}

set.seed(42)

rnorm(n = 50, m = 0, sd = 1) |> 
  hist()

```

## Generate two normally distributed data sets and fit a *t*-test

Note the population means are equivalent. What does a *t*-test test for?

```{r}

set.seed(42)

t.test(rnorm(n = 50, m = 0, sd = 1), 
       rnorm(n = 50, m = 0, sd = 1))

```

## Generate two normally distributed data sets and fit a *t*-test and extract the *p*-value

```{r}

set.seed(42)

t.test(rnorm(n = 50, m = 0, sd = 1), rnorm(n = 50, m = 0, sd = 1))$p.value

```

## Do this generate-analyze-extract many times

Using the `replicate()` function, and plot the *p*-values using `hist()`.

```{r}

set.seed(42)

hist(replicate(1000, t.test(rnorm(n = 50, m = 0, sd = 1), rnorm(n = 50, m = 0, sd = 1))$p.value))

```

## Increase the number of iterations

Why is the distribution more uniform?

```{r}

set.seed(42)

hist(replicate(100000, t.test(rnorm(n = 50, m = 0, sd = 1), rnorm(n = 50, m = 0, sd = 1))$p.value))

```

## Compare with distribution of *p*-values when population means are not equal

'Small', 'medium' and 'large' differences in population means.

```{r}

set.seed(42)

hist(replicate(100000, t.test(rnorm(n = 50, m = 0, sd = 1), rnorm(n = 50, m = 0.2, sd = 1))$p.value))
hist(replicate(100000, t.test(rnorm(n = 50, m = 0, sd = 1), rnorm(n = 50, m = 0.5, sd = 1))$p.value))
hist(replicate(100000, t.test(rnorm(n = 50, m = 0, sd = 1), rnorm(n = 50, m = 0.8, sd = 1))$p.value))

```

## Check your learning

#### What are the five core components of a Monte Carlo simulation?

::: {.callout-note collapse="true" title="Click to show answer"}
1. Construct an experiment
2. Generate data
3. Analyze data
4. Repeat these many times
5. Summarize results over the experiment conditions
:::

#### Code implementing each 

Can you mentally match these five components onto the above code (sometimes in different chunks!) to understand how, together, they implement a simulation that answers the question "what is the distribution of *p*-values under the null vs. alternative hypothesis?"


