---
title: "Simulation template - simulation name goes here"
author: "Authors names go here"
date: today
format:
  html:
    toc: true
    toc_float: true
    code-fold: false
    code-overflow: wrap
    code-tools: true
    warning: false
    df-print: kable
    theme:
      light: cosmo
      dark: darkly
    highlight-style:
      light: breeze
      dark: dracula
    include-in-header:
      text: |
        <style>
          .kable-table table { width: 70%; margin: 0 auto; }
          .quarto-color-scheme-toggle::after { content: " Dark mode"; font-family: sans-serif; font-size: 0.85em; }
        </style>
---

```{r, include=FALSE}
# settings: hidden from rendered file
# disable messages and warnings in rendered files
knitr::opts_chunk$set(message = FALSE,
                      warning = FALSE)

# disable scientific numbers in output
options(scipen=999)
```

# Introduction

<!-- Specify the simulation that will be run here. -->

## Project description

<!-- Provide a brief description of the project, including any empirical examples. -->



## Prior related work

<!-- Did any contributors already conduct related simulation studies on this specific question, including any preliminary simulations? -->



## Aims

<!-- What is the aim of the simulation study? Be specific about the statistical task (estimation, hypothesis testing, model selection, prediction, design). -->



## Data-Generating Process

### DGP specification approach

<!-- How will the parameters for the DGP be specified? (parametric based on real data / parametric / resampled) If based on real data, describe the dataset and model used. -->



### DGP factors

<!-- What will be the different factors of the DGP — i.e., parameters/settings varied across simulation conditions? -->



### Factor values and settings

<!-- Provide specific factor values and any settings held constant across conditions. Include justification for chosen values where relevant. -->



### Factor combination and number of conditions

<!-- If there is more than one factor: how will factor levels be combined? (fully factorial / partially factorial / one-at-a-time / scattershot) How many conditions does this create? -->



## Estimands and Targets

<!-- What are the estimands and/or targets of the simulation study? Distinguish primary and secondary targets if applicable. -->



## Methods and extracted quantities

<!-- How many and which methods will be included? What quantities will be extracted from each? Include code or pseudocode where helpful. Justify method and parameter choices; note where package defaults are used. -->



## Performance and Uncertainty

### Performance measures

<!-- Which performance measures will be used? Provide formulas and justification. If multiple parameters are aggregated, specify how. -->



### Monte Carlo uncertainty

<!-- How will Monte Carlo Standard Errors (MCSEs) be calculated and reported? -->

<!-- **Example1:** We will report Monte Carlo uncertainty in tables (MCSEs next to the estimated performance measures) and in plots (error bars with ±1 MCSE around estimated performance measures). We will use the formulas provided in Siepe et al. (2024) to calculate MCSEs. -->
<!-- **Example2:** We will report Monte Carlo uncertainty in tables (Jackknife MCSEs next to the estimated performance measures) and in plots (error bars with ±1 Jackknife MCSE around estimated performance measures). We will use the Jackknife-based MCSEs computed using the [rsimsum (Gasparini 2018) / simhelpers (Joshi and Pustejovsky 2022)] R package. -->



### Number of simulation repetitions

<!-- How many repetitions per condition? Justify the number (sample size calculation, computational constraints, rule of thumb, etc.). -->

<!-- **Example1:** We will perform 10,000 repetitions per condition. We determined this number by aiming for a MCSE of 0.005 for the type I error rate and the power under the “worst-case” rejection rate of 50% in the sense that the MCSE is maximal for a given number of repetitions (0.50×((1-0.50))⁄〖0.005〗^2 =10,000 repetitions). -->
<!-- **Example2:** We determined the required number of repetitions to achieve a MCSE of 0.005 for the bias for each of the methods. The sample size calculation requires the empirical variance of the effect estimates S_θ ^^2 for each method. Since the empirical variances of the effect estimates can vary across simulation conditions, we compute the sample size using the largest estimated variance across all conditions. We obtain the empirical variance estimates for each condition and method using 100 pilot simulation runs. We found that the required sample sizes would be (S_θ ^^2)⁄〖0.005〗^2 =1,986 for ANCOVA, (S_θ ^^2)⁄〖0.005〗^2 =3,812 for change score analysis, and (S_θ ^^2)⁄〖0.005〗^2 =1,996 for post score analysis. -->



### Non-convergence and missing values

<!-- How will missing values due to non-convergence or other failures be handled? -->

<!-- **Example:** We do not expect missing values or non-convergence. If we observe any non-convergence, we exclude the non-converged cases case-wise (keeping the converged values from the other methods in the same repetition) and report the number of non-converged cases per method and condition. -->



### Interpretation of performance measures *(optional)*

<!-- What constitutes acceptable/unacceptable performance? Are any regression models or effect sizes planned for analysing simulation results? -->





Experimentally manipulated:

The simulation varies the number of participants sampled (N per condition = 25 to 200 in 25s). 

Static:

- Balanced N in each condition.
- Population difference in means ($\Delta\mu$) = 0.
- Population standard deviation ($\sigma$) in each condition (equal in both conditions) = 1.
- Population standardized effect size is therefore = 0 ($\Delta\mu / \sigma_{pooled} = 0$).
- Single analytic strategy: Students' t-test testing differences in means from zero ($\Delta\mu = 0$), alpha = 5%.

# Methods

## Dependencies

```{r}
library(tidyr)
library(dplyr)
library(furrr)
library(janitor)
library(parameters)
library(simhelpers)
library(ggplot2)
library(scales)
library(knitr)
library(kableExtra)
library(janitor)

# set up parallelisation for {furrr}
plan(multisession)
```

## Functions 

### Data generating process

```{r}
generate_data <- function(n_per_condition,
                          mean_control,
                          mean_intervention,
                          sd) {
  
  data_control <- 
    tibble(condition = "control",
           score = rnorm(n = n_per_condition, mean = mean_control, sd = sd))
  
  data_intervention <- 
    tibble(condition = "intervention",
           score = rnorm(n = n_per_condition, mean = mean_intervention, sd = sd))
  
  # combine
  data <- bind_rows(data_control,
                    data_intervention) |>
    # ensure control is the reference condition
    mutate(condition = factor(condition, levels = c("intervention", "control")))
  
  return(data)
}
```

### Analysis

```{r}
analyse <- function(data, alpha = 0.05) {
  # fit Students' t-test
  fit <- t.test(formula = score ~ condition,
                data = data,
                var.equal = TRUE,
                conf.level = 1 - alpha,
                alternative = "two.sided")
  
  # extract p value
  results <- fit %>%
    model_parameters() %>%
    as_tibble() %>% 
    clean_names() %>% 
    # select columns of interest
    select(mean_diff = difference,
           ci_lower = ci_low,
           ci_upper = ci_high,
           p)
    
  return(results)
}
```

## Define experiment

```{r}
experiment_parameters_grid <- expand_grid(
  n_per_condition = seq(from = 10, to = 200, by = 10),
  mean_control = 0,
  mean_intervention = c(0, 0.2, 0.5, 0.8),
  sd = 1,
  iteration = 1:100L
) |>
  # define population values
  mutate(population_mean_diff = mean_intervention - mean_control) |>
  # define unique conditions
  mutate(condition = paste0("N = ", n_per_condition,
                            ", Pop mean diff = ", population_mean_diff))
```

## Run simulation

```{r}
set.seed(42)

simulation <- experiment_parameters_grid |>
  # generate data
  mutate(data = future_pmap(.l = list(n_per_condition = n_per_condition, 
                                      mean_control = mean_control, 
                                      mean_intervention = mean_intervention, 
                                      sd = sd),
                            .f = generate_data,
                            .progress = TRUE,
                            .options = furrr_options(seed = TRUE))) |>
  # analyse data
  mutate(results = future_pmap(.l = list(data = data),
                               .f = analyse,
                               .progress = TRUE,
                               .options = furrr_options(seed = TRUE)))

# optionally save to disk
#write_rds(x = simulation, file = "simulation.rds", compress = "gz")
```

# Results

<!-- Summarize over iterations by condition -->
<!-- Performance and uncertainty of the method(s) across conditions -->

```{r}
simulation_summary <- simulation |>
  unnest(results) |>
  group_by(n_per_condition,
           population_mean_diff,
           condition) |>
  # performance and uncertainty metrics for inference (hypothesis tests with binary decisions)
  reframe(
    calc_rejection(data       = pick(everything()),
                   p_values   = p, 
                   alpha      = 0.05)
  ) |>
  rename(empirical_detection_rate = rej_rate,
         empirical_detection_rate_mcse = rej_rate_mcse)

# # identical summarize results, more manual:
#   summarize(
#     empirical_detection_rate  = mean(test_result),
#     empirical_detection_rate_mcse = sqrt(empirical_detection_rate * (1 - empirical_detection_rate) / n()),
#     .groups = "drop"
#   )
```

### Table

```{r}
simulation_summary |>
  select(n_per_condition, 
         population_mean_diff,
         empirical_detection_rate, 
         empirical_detection_rate_mcse) |>
  mutate_if(is.numeric, janitor::round_half_up, digits = 3) |>
  kable() |>
  kable_classic(full_width = FALSE)
```

### Plot

```{r}
ggplot(simulation_summary, 
       aes(x = n_per_condition, 
           y = empirical_detection_rate, 
           color = as.factor(population_mean_diff))) +
  geom_hline(yintercept = 0.05, linetype = "dashed") +
  geom_linerange(aes(ymin = empirical_detection_rate - empirical_detection_rate_mcse,
                     ymax = empirical_detection_rate + empirical_detection_rate_mcse),
                 color = "black") +
  geom_line() +
  geom_point() +
  scale_x_continuous(name = "N per condition",
                     breaks = breaks_pretty(n = 8)) +
  scale_y_continuous(name = "Empirical detection rate\nof Student's t test p-values", 
                     limits = c(0, 1),
                     breaks = c(0, .05, .25, .5, .75, 1)) +
  theme_linedraw() +
  theme(panel.grid.minor = element_blank()) +
  guides(color = guide_legend(title = "Population mean difference", 
                              reverse = TRUE))
```

### Table

```{r}
simulation_summary_estimation |>
  select(n_per_condition, 
         population_mean_diff,
         bias, 
         bias_mcse,
         var,
         var_mcse) |>
  mutate_if(is.numeric, janitor::round_half_up, digits = 3) |>
  kable() |>
  kable_classic(full_width = FALSE)
```

### Plot

```{r}
ggplot(simulation_summary_estimation, 
       aes(x = n_per_condition, 
           y = bias, 
           color = as.factor(population_mean_diff))) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_linerange(aes(ymin = bias - bias_mcse,
                     ymax = bias + bias_mcse),
                 color = "black") +
  geom_line() +
  geom_point() +
  scale_x_continuous(name = "N per condition",
                     breaks = breaks_pretty(n = 8)) +
  scale_y_continuous(name = "Bias in estimation\nof population mean difference", 
                     limits = c(-1, 1),
                     breaks = breaks_pretty(n = 8)) +
  theme_linedraw() +
  theme(panel.grid.minor = element_blank()) +
  guides(color = guide_legend(title = "Population mean difference"))
```

### Table

```{r}
simulation_summary_coverage |>
  select(n_per_condition, 
         population_mean_diff,
         coverage, 
         coverage_mcse,
         width,
         width_mcse) |>
  mutate_if(is.numeric, janitor::round_half_up, digits = 3) |>
  kable() |>
  kable_classic(full_width = FALSE)
```

### Plot

```{r}
ggplot(simulation_summary_coverage, 
       aes(x = n_per_condition, 
           y = coverage, 
           color = as.factor(population_mean_diff))) +
  geom_hline(yintercept = 0.95, linetype = "dashed") +
  geom_linerange(aes(ymin = coverage - coverage_mcse,
                     ymax = coverage + coverage_mcse),
                 color = "black") +
  geom_line() +
  geom_point() +
  scale_x_continuous(name = "N per condition",
                     breaks = breaks_pretty(n = 8)) +
  scale_y_continuous(name = "Coverage of 95% CIs\nfor population mean difference", 
                     limits = c(0, 1),
                     breaks = c(0, .25, .5, .75, .95, 1)) +
  theme_linedraw() +
  theme(panel.grid.minor = element_blank()) +
  guides(color = guide_legend(title = "Population mean difference"))

ggplot(simulation_summary_coverage, 
       aes(x = n_per_condition, 
           y = width, 
           color = as.factor(population_mean_diff))) +
  geom_linerange(aes(ymin = width - width_mcse,
                     ymax = width + width_mcse),
                 color = "black") +
  geom_line() +
  geom_point() +
  scale_x_continuous(name = "N per condition",
                     breaks = breaks_pretty(n = 8)) +
  scale_y_continuous(name = "Width of 95% CIs", 
                     breaks = breaks_pretty(n = 8)) +
  theme_linedraw() +
  theme(panel.grid.minor = element_blank()) +
  guides(color = guide_legend(title = "Population mean difference"))
```

# Discussion

<!-- Discuss and interpret results of the simulation here: what was the performance of the method(s) across the conditions? What can we conclude with regard to the aims? -->

# Session info

<!-- Recorded for reproducibility -->

```{r}
sessionInfo()
```




