---
title: "Simulation template: Cohen's d is a biased estimator"
#author: "Authors names go here"
date: today
format:
  html:
    embed-resources: true # ignored when rendering book
    toc: true
    toc_float: true
    code-fold: false
    code-overflow: wrap
    code-tools: true
    warning: false
    df-print: kable
    theme:
      light: cosmo
      dark: darkly
    highlight-style:
      light: breeze
      dark: dracula
    include-in-header:
      text: |
        <style>
          .kable-table table { width: 70%; margin: 0 auto; }
          .quarto-color-scheme-toggle::after { content: " Dark mode"; font-family: sans-serif; font-size: 0.85em; }
        </style>
---

This modifies the lego simulation to examine the bias in Cohen's d in small sample sizes, and the correction of this bias via Hedge's g (the J correction).

This simulation serves the second purpose of demonstrating that the experiments implemented in Monte Carlo simulations often contain *within* factors, not just between. I.e., in this case, the analysis of each iteration's dataset two different ways (with and without Hedge's correction), rather than comparing different datasets analyzed different ways.

As such, some factors in the experiment are *not* in the expand_grid, but are instead a feature of the analysis function or the piped workflow in the running of the simulation (e.g., calling two variations of the analysis to create different results columns).

```{r, include=FALSE}
# settings: hidden from rendered file
# disable messages and warnings in rendered files
knitr::opts_chunk$set(message = FALSE,
                      warning = FALSE)

# disable scientific numbers in output
options(scipen=999)
```

# Methods

## Setup

```{r}
# dependencies
library(tidyr)
library(dplyr)
library(readr)
library(furrr)
library(effsize)
library(simhelpers)
library(ggplot2)
library(scales)
library(knitr)
library(kableExtra)

# set up parallelisation for {furrr}
# `furrr_options(seed = TRUE)` (used below in future_pmap calls) advances an L'Ecuyer-CMRG stream rather than re-seeding per worker, giving the parallel-safe reproducibility recommended in Siepe et al. (2024, p. 8).
# use availableCores to define the number of cores in case sim is being run on a HPC that might have more physical cores than are available to the session
plan(multisession, workers = parallelly::availableCores())
```

## Functions 

### Data generating process

```{r}
generate_data <- function(mean_control,
                          mean_intervention,
                          sd_control,
                          sd_intervention,
                          n_control,
                          n_intervention) {
  
  data_control <- 
    tibble(condition = "control",
           score = rnorm(n = n_control, mean = mean_control, sd = sd_control))
  
  data_intervention <- 
    tibble(condition = "intervention",
           score = rnorm(n = n_intervention, mean = mean_intervention, sd = sd_intervention))
  
  # combine
  data <- bind_rows(data_control,
                    data_intervention) |>
    # ensure control is the reference condition
    mutate(condition = factor(condition, levels = c("intervention", "control")))
  
  return(data)
}
```

### Analysis

```{r}
analyze <- function(data, alpha = 0.05) {
  res_t_test <- t.test(formula = score ~ condition,
                       data = data,
                       var.equal = TRUE,
                       alternative = "two.sided",
                       conf.level = 1-alpha)

  # both estimators are calculated from the *same* data set, so any difference
  # between them is due to Hedges' small sample correction rather than to sampling
  # variation. this is much more efficient than treating the correction as another
  # factor in the experiment grid, which would compare estimates calculated from
  # different data sets and therefore bury the correction under Monte Carlo noise.
  res_cohens_d <- effsize::cohen.d(formula = score ~ condition,
                                   data = data,
                                   pooled = TRUE,
                                   hedges.correction = FALSE,
                                   conf.level = 1-alpha)

  res_hedges_g <- effsize::cohen.d(formula = score ~ condition,
                                   data = data,
                                   pooled = TRUE,
                                   hedges.correction = TRUE,
                                   conf.level = 1-alpha)

  p <- res_t_test$p.value

  tibble(
    mean_diff          = unname(res_t_test$estimate[1] - res_t_test$estimate[2]),
    mean_diff_ci_lower = res_t_test$conf.int[1],
    mean_diff_ci_upper = res_t_test$conf.int[2],
    cohens_d           = res_cohens_d$estimate,
    cohens_d_ci_lower  = res_cohens_d$conf.int[["lower"]],
    cohens_d_ci_upper  = res_cohens_d$conf.int[["upper"]],
    hedges_g           = res_hedges_g$estimate,
    hedges_g_ci_lower  = res_hedges_g$conf.int[["lower"]],
    hedges_g_ci_upper  = res_hedges_g$conf.int[["upper"]],
    p                  = p
    # note: no `significant` column. it would be a copy of `p < alpha` stored for every
    # iteration, and can always be recalculated from `p` when it is needed
  )
}
```

### Helper function

By not saving the data to a data column in the data frame, we can speed up the simulation a lot and produce a smaller data frame. This can matter when they get more complicated.

```{r}
run_one_iteration <- function(mean_control,
                              mean_intervention,
                              sd_control,
                              sd_intervention,
                              n_control,
                              n_intervention,
                              alpha = 0.05) {

  data <- generate_data(mean_control      = mean_control,
                        mean_intervention = mean_intervention,
                        sd_control        = sd_control,
                        sd_intervention   = sd_intervention,
                        n_control         = n_control,
                        n_intervention    = n_intervention)

  results <- analyze(data, alpha = alpha)

  return(results)
}
```

## Define experiment

```{r}
experiment_parameters_grid <- 
  expand_grid(
    mean_control = 0,
    mean_intervention = c(0, 0.2, 0.5, 0.8), # when both sd=1, diff between m-ctrl and m-intr = SMD
    sd_control = 1,
    n_control = round(seq(from = 5, to = 50, by = 5), 0), # round because seq can accrue floating point errors
    alpha = 0.05,
    # note that the Hedges' correction is *not* a factor in the grid: both estimators
    # are calculated from every simulated data set inside analyze() instead
    iteration = 1:10000L
  ) |>
  
  # make ns per condition equal without crossing these factors in the grid
  mutate(n_intervention = n_control,
         sd_intervention = sd_control,
         n_total = n_control + n_intervention) |>
  
  # define population values
  mutate(
    population_mean_diff = mean_intervention - mean_control,
    population_sd_pooled = sqrt(
      ((n_intervention - 1) * sd_intervention^2 +
         (n_control - 1) * sd_control^2) /
        (n_intervention + n_control - 2)
    ),
    population_smd = population_mean_diff / population_sd_pooled
  )
```

## Run simulation

Data are generated and analyzed inside `run_one_iteration()`, so only the results are returned and nothing but the results is stored.

```{r}
dir.create("results", showWarnings = FALSE)

if (file.exists("results/simulation_cohens_d_biased.rds")) {
  simulation <- read_rds("results/simulation_cohens_d_biased.rds")
} else {
  set.seed(42)

  simulation <- experiment_parameters_grid |>
    # shuffle the rows of the grid to load balance the computationally expensive iterations evenly across workers
    slice_sample(prop = 1) |>
    
    # generate + analyse in one step via the helper
    mutate(results = future_pmap(
      .l = list(mean_control      = mean_control,
                mean_intervention = mean_intervention,
                sd_control        = sd_control,
                sd_intervention   = sd_intervention,
                n_control         = n_control,
                n_intervention    = n_intervention,
                alpha             = alpha),
      .f = run_one_iteration,
      .progress = TRUE,
      .options  = furrr_options(seed = TRUE)
    )) |>
    unnest(results)

  write_rds(x = simulation, file = "results/simulation_cohens_d_biased.rds", compress = "gz")
}
```

# Results

Each condition is summarized over its iterations. Because every iteration produced all three estimates, the summary has one set of metrics per estimator, distinguished by a column prefix: `mean_diff_`, `cohens_d_`, and `hedges_g_`.

```{r}
simulation_summary <- simulation |>
  group_by(
    # group by all factors in the expand_grid other than iterations!
    mean_control,
    mean_intervention,
    sd_control,
    sd_intervention,
    n_control,
    n_intervention,
    alpha,
    # variables created from the above parameters, to keep them in the output:
    n_total,
    population_mean_diff,
    population_sd_pooled,
    population_smd
  ) |>
  reframe(
    # performance and uncertainty metrics for inference (hypothesis tests with binary decisions)
    calc_rejection(data       = pick(everything()),
                   p_values   = p,
                   alpha      = 0.05),
    # performance and uncertainty metrics for estimation (continuous estimates):
    # bias (systematic error) and the empirical standard error (precision) of the point
    # estimate. calc_absolute() always returns columns named bias/stddev, so each call is
    # prefixed to say which estimator it refers to and to avoid name collisions.
    # note the different targets: the mean difference is estimated in raw units, the two
    # standardized effect sizes in standard deviation units
    calc_absolute(data        = pick(everything()),
                  estimates   = mean_diff,
                  true_param  = population_mean_diff,
                  criteria    = c("bias", "stddev")) |>
      rename_with(\(x) paste0("mean_diff_", x)),
    calc_absolute(data        = pick(everything()),
                  estimates   = cohens_d,
                  true_param  = population_smd,
                  criteria    = c("bias", "stddev")) |>
      rename_with(\(x) paste0("cohens_d_", x)),
    calc_absolute(data        = pick(everything()),
                  estimates   = hedges_g,
                  true_param  = population_smd,
                  criteria    = c("bias", "stddev")) |>
      rename_with(\(x) paste0("hedges_g_", x)),
    # calibration (coverage) and precision (width) of the interval estimates
    calc_coverage(data        = pick(everything()),
                  lower_bound = mean_diff_ci_lower,
                  upper_bound = mean_diff_ci_upper,
                  true_param  = population_mean_diff,
                  criteria    = c("coverage", "width")) |>
      rename_with(\(x) paste0("mean_diff_", x)),
    calc_coverage(data        = pick(everything()),
                  lower_bound = cohens_d_ci_lower,
                  upper_bound = cohens_d_ci_upper,
                  true_param  = population_smd,
                  criteria    = c("coverage", "width")) |>
      rename_with(\(x) paste0("cohens_d_", x)),
    calc_coverage(data        = pick(everything()),
                  lower_bound = hedges_g_ci_lower,
                  upper_bound = hedges_g_ci_upper,
                  true_param  = population_smd,
                  criteria    = c("coverage", "width")) |>
      rename_with(\(x) paste0("hedges_g_", x)),
    # 95% prediction intervals: the 2.5th and 97.5th percentiles of the estimates
    # across iterations, i.e. the range of results a single new study would be
    # expected to produce 95% of the time. simhelpers has no function for this, so
    # it is computed directly. note that this is a very different quantity to the
    # 95% CI widths above: the CI width is a property of one study's interval,
    # whereas this is the spread of the point estimates themselves. If the CIs are
    # well calibrated, the mean CI width should be similar to the width of the
    # prediction interval.
    tibble(
      mean_diff_estimate = mean(mean_diff),
      mean_diff_pi_lower = unname(quantile(mean_diff, probs = 0.025)),
      mean_diff_pi_upper = unname(quantile(mean_diff, probs = 0.975)),
      mean_diff_pi_width = mean_diff_pi_upper - mean_diff_pi_lower,
      cohens_d_estimate  = mean(cohens_d),
      cohens_d_pi_lower  = unname(quantile(cohens_d, probs = 0.025)),
      cohens_d_pi_upper  = unname(quantile(cohens_d, probs = 0.975)),
      cohens_d_pi_width  = cohens_d_pi_upper - cohens_d_pi_lower,
      hedges_g_estimate  = mean(hedges_g),
      hedges_g_pi_lower  = unname(quantile(hedges_g, probs = 0.025)),
      hedges_g_pi_upper  = unname(quantile(hedges_g, probs = 0.975)),
      hedges_g_pi_width  = hedges_g_pi_upper - hedges_g_pi_lower
    )
  ) |>
  rename(empirical_detection_rate = rej_rate,
         empirical_detection_rate_mcse = rej_rate_mcse,
         # calc_absolute() calls its SD of the estimates "stddev"; it is the
         # empirical standard error of the estimator across iterations
         mean_diff_empirical_se = mean_diff_stddev,
         mean_diff_empirical_se_mcse = mean_diff_stddev_mcse,
         cohens_d_empirical_se = cohens_d_stddev,
         cohens_d_empirical_se_mcse = cohens_d_stddev_mcse,
         hedges_g_empirical_se = hedges_g_stddev,
         hedges_g_empirical_se_mcse = hedges_g_stddev_mcse)
```

The summary above is wide: one row per condition, with one set of columns per estimator. Tables and plots that compare the estimators are easier to write from a long version, with one row per condition *per estimator*. `names_to = c("estimator", ".value")` splits each column name at the prefix, so `cohens_d_bias` and `hedges_g_bias` both become a `bias` column, distinguished by the new `estimator` column.

```{r}
simulation_summary_long <- simulation_summary |>
  pivot_longer(cols = matches("^(mean_diff|cohens_d|hedges_g)_"),
               names_to = c("estimator", ".value"),
               names_pattern = "^(mean_diff|cohens_d|hedges_g)_(.*)$") |>
  # factor() relabels and orders the estimators in one step
  mutate(estimator = factor(estimator,
                            levels = c("mean_diff", "cohens_d", "hedges_g"),
                            labels = c("Mean difference", "Cohen's d", "Hedges' g")))
```

### Table

```{r}
# two ways of writing a table cell: a point estimate with its MCSE in parentheses,
# or an interval in square brackets
mcse_cell     <- function(estimate, mcse) paste0(estimate, " (", mcse, ")")
interval_cell <- function(lower, upper)   paste0("[", lower, ", ", upper, "]")

# helper to avoid repeating the same wide-table code for each performance metric.
# `cell` is an expression built from the long summary's columns, e.g. mcse_cell(bias, bias_mcse).
# it is called on `simulation_summary_long` filtered to the estimator(s) of interest, and
# each row of the table is a combination of sample size and estimator.
wide_table <- function(data, cell, caption, header = "Population mean difference",
                       note = "Values in parentheses are ±1 Monte Carlo Standard Error.") {
  data |>
    arrange(n_control, estimator) |>
    mutate(`n per condition` = as.character(n_control),
           population_mean_diff = as.character(population_mean_diff)) |>
    mutate(across(where(is.numeric), \(x) scales::number(x, accuracy = 0.001))) |>
    mutate(cell_string = {{ cell }}) |>
    select(`n per condition`,
           estimator,
           population_mean_diff,
           cell_string) |>
    pivot_wider(names_from = population_mean_diff,
                values_from = cell_string) |>
    kable(caption = caption) |>
    kable_styling(full_width = FALSE) |>
    # setNames() because the header label is a variable: the first two columns are
    # the row labels (sample size and estimator), the next four the effect sizes
    add_header_above(setNames(c(2, 4), c(" ", header))) |>
    footnote(general = note,
             general_title = "Note.",
             footnote_as_chunk = TRUE)
}

# better wider tables
# bias: systematic error in the estimate of the population mean difference.
# the mean difference is in raw units and is not affected by Hedges' correction, so it
# serves as a reference point: an estimator that is unbiased at every sample size
wide_table(filter(simulation_summary_long, estimator == "Mean difference"),
           mcse_cell(bias, bias_mcse),
           caption = "Bias (±1 MCSE) of the estimated mean difference by sample size and population mean difference")

# bias: systematic error in the estimate of the population standardized mean difference.
# this is the comparison the simulation was built for: Cohen's d overestimates the
# population standardized mean difference, especially at small samples, whereas
# Hedges' g corrects for this
wide_table(filter(simulation_summary_long, estimator %in% c("Cohen's d", "Hedges' g")),
           mcse_cell(bias, bias_mcse),
           caption = "Bias (±1 MCSE) of the estimated standardized mean difference by sample size, estimator, and population standardized mean difference",
           header = "Population standardized mean difference")
```

### Plot

```{r}
# helper to avoid repeating the same plotting code for each performance metric.
# it is called on `simulation_summary_long`, so the estimator is a column and can be
# mapped to color: each population effect size gets its own facet, which keeps a single
# panel from having to carry both the effect sizes and the estimators at once.
# `lower` and `upper` are expressions built from the long summary's columns: ±1 MCSE for
# the performance metrics (matching the values in parentheses in the tables above),
# or the bounds of the prediction interval for the plots of the estimates themselves.
metric_plot <- function(data,
                        estimate,
                        lower,
                        upper,
                        y_name,
                        facet_var      = population_smd,
                        facet_name     = "Population SMD",
                        reference_line = NULL,
                        y_limits       = NULL,
                        y_breaks       = breaks_pretty(n = 8)) {

  # pasted into a single column so the facet strips are self describing
  data <- data |>
    mutate(facet_label = paste0(facet_name, " = ", {{ facet_var }}))

  p <- ggplot(data,
              aes(x = n_control,
                  y = {{ estimate }},
                  linetype = estimator,
                  color = estimator)) +
    scale_color_viridis_d(begin = 0.2, end = 0.8)

  # e.g. the nominal alpha for detection rates under the null, or zero for bias
  if (!is.null(reference_line)) {
    p <- p + geom_hline(yintercept = reference_line, linetype = "dashed")
  }

  p +
    geom_linerange(aes(ymin = {{ lower }},
                       ymax = {{ upper }}),
                   color = "black") +
    geom_line() +
    geom_point() +
    facet_wrap(~ facet_label) +
    scale_x_continuous(name = "N per condition",
                       breaks = breaks_pretty(n = 5)) +
    scale_y_continuous(name = y_name,
                       limits = y_limits,
                       breaks = y_breaks) +
    theme_linedraw() +
    theme(panel.grid.minor = element_blank(),
          legend.position = "bottom") +
    guides(linetype = guide_legend(title = "Estimator"),
           color = guide_legend(title = "Estimator"))
}

# bias: systematic error in the estimate of the population mean difference.
# note that the y axis is not fixed to a wide range here: the bias being demonstrated
# is a few hundredths of a standard deviation, so a c(-1, 1) axis would hide it entirely
metric_plot(filter(simulation_summary_long, estimator == "Mean difference"),
            estimate       = bias,
            lower          = bias - bias_mcse,
            upper          = bias + bias_mcse,
            y_name         = "Bias in estimation\nof population mean difference",
            facet_var      = population_mean_diff,
            facet_name     = "Population mean difference",
            reference_line = 0) +
  coord_cartesian(ylim = c(-0.1, 0.1))

# bias: systematic error in the estimate of the population standardized mean difference.
# Cohen's d sits above zero at small sample sizes and converges on it as N grows,
# whereas Hedges' g is on zero throughout
metric_plot(filter(simulation_summary_long, estimator %in% c("Cohen's d", "Hedges' g")),
            estimate       = bias,
            lower          = bias - bias_mcse,
            upper          = bias + bias_mcse,
            y_name         = "Bias in estimation\nof population standardized mean difference",
            reference_line = 0)
```

# References

Hedges, L. V. (1981). Distribution Theory for Glass’s Estimator of Effect size and Related Estimators. *Journal of Educational Statistics*, 6(2), 107-128. doi: [10.3102/10769986006002107](https://doi.org/10.3102/10769986006002107)

Hussey, I. & Cummins, J. (2026). *Understanding Statistics through Monte Carlo simulations*. [simulations.tidyver.se](https://simulations.tidyver.se) doi: [10.5281/zenodo.19420665](https://doi.org/10.5281/zenodo.19420665)

Miratrix and Pustejovsky (2026). *Designing Monte Carlo Simulations in R*. [jepusto.github.io/Designing-Simulations-in-R](https://jepusto.github.io/Designing-Simulations-in-R/)

Morris, T. P., White, I. R., & Crowther, M. J. (2019). Using simulation studies to evaluate statistical methods. *Statistics in Medicine*, 38(11), 2074–2102. https://doi.org/10.1002/sim.8086

Siepe, B. S., Bartoš, F., Morris, T. P., Boulesteix, A.-L., Heck, D. W., & Pawel, S. (2024). Simulation studies for methodological research in psychology: A standardized template for planning, preregistration, and reporting. *Psychological Methods*. https://doi.org/10.1037/met0000695

Strobl, C. et al. (2024). *Simulationsstudien in R: Design und praktische Durchführung*. doi: [doi.org/10.1007/978-3-662-70561-2](https://doi.org/10.1007/978-3-662-70561-2) [English translation should be available in late 2026]

# Session info

```{r}
sessionInfo()
```

